---
title: "20250521-CSP_CFP literature review"
date: today
dateformat: long
author: 
  - name: Lisa Sheenan
    email: lisa.sheenan@ucd.ie
    affiliations:
      - id: ucd
        name: University College Dublin
        department: UCD College of Business
        address: Dublin, Ireland
    attributes:
      corresponding: true  
  - name: Ying (Veronica) Zhang
    email: veronica.zhang@qub.ac.uk
    affiliations:
      - id: qbs
        name: Queen's University, Belfast
        department: Queen's Business School
        address: Belfast, UK
  - name: Atul Saxena
    email: asaxena@ggc.edu
    affiliations:
      - id: ggc
        name: Georgia Gwinnett College
        address: Georgia, US

abstract: |
  The following sections review the theoretical frameworks of the relationship between ESG reporting and financial institutions. The theoretical frameworks include stakeholder theory, legitimacy theory, and institutional theory. The study also reviews emerging frameworks such as natural capital and double materiality. The review proposes a Stakeholder-Legitimacy Framework, integrating stakeholder theory, legitimacy theory, and institutional isomorphism to explain how banks’ ESG disclosures emerge from dynamic negotiations between external pressures (e.g., regulators, investors) and strategic legitimacy-seeking behaviors. Key findings from the review of empirical research reveal that while ESG disclosures mitigate bank-specific risks, conflicting evidence persists on their financial performance impacts. Critical gaps include underexplored transmission effects of regulations through lending networks, inconsistent roles of institutional and state ownership, and tensions between symbolic compliance and substantive action.
  The research question can be: What is the relationship between ESG reporting and banks? How does ESG reporting regulation impact bank performance, for example, the EBA's Green Asset Ratios? Using the integrated stakeholder-legitimacy framework, we can examine the influencing factors such as regulation and bank performance.    

keywords: 
  - ESG disclosure
  - IFRS S1 and S2
  - Stakeholder-Legitimacy framework
  - bank risk
  - ownership
  - climate change

bibliography: review.bib
format: 
  elsevier-pdf: 
    keep-tex: true
    header-includes: |
      \usepackage{graphicx}
      \usepackage{adjustbox}
    journal:
      name: "Corporate Governance: An International Review"
      cite-style: authoryear
execute: 
  echo: false
  eval: true
  output: true
  warning: false
  error: false
  include: true
---

```{r set up}
# label: libraries
# output: false
# include: false

library(tidyverse)
library(kableExtra)
library(DiagrammeR)
library(DiagrammeRsvg)
library(rsvg)
library(knitr)
library(tinytex)

```


# ESG disclsoure and Banks

## Theoretical Foundation

The relationship between environmental, social, and governance (ESG) reporting and financial institutions has garnered increasing scholarly attention, driven by global sustainability challenges. This section synthesizes theoretical frameworks that explain how and why ESG reporting influences financial institutions’ operations, decision-making, and market outcomes. Four interconnected theoretical streams provide the foundation: organizational theories, economic theories, institutional theory, and emerging frameworks such as natural capital and double materiality.

### Organizational Theories

Organizational theories, particularly stakeholder theory and legitimacy theory, highlight the institutional motivations and external pressures driving ESG reporting practices.  Stakeholder theory, developed by @FREEMAN1984, posits that organizations are accountable to a broad range of stakeholders beyond shareholders, including regulators, customers (investors), employees, and civil society. Building on this, @MITCHELL1997 proposed the theory of stakeholder salience emphasizing understanding why and how managers prioritize certain stakeholders, rather than prescribing whom they should prioritize. Their theory provides tools to identify stakeholder types via a typology and explain managerial responses through the concept of salience, which assesses stakeholders' legitimacy, power, and urgency. This typology offers a framework for financial institutions to navigate competing stakeholder interests effectively. @DONALDSON1995 provided theoretical justification for stakeholder theory, emphasizing its descriptive, instrumental, and normative validity in explaining corporate behavior. @ROBERTS1992 conducts an application of stakeholder theory to explain corporate social responsibility disclosure. The study finds that measures of stakeholder power, strategic posture, and economic performance are significantly associated with levels of corporate social disclosure. Banks' ESG disclosure, for instance, on climate risks or green lending portfolios are often framed as responses to investor demands for transparency or regulatory mandates like the EU’s Corporate Sustainability Reporting Directive (CSRD). 

Legitimacy theory, foundationally developed by @SUCHMAN1995, identifies three primary forms of legitimacy[^03]—pragmatic, moral, and cognitive—that organizations seek to maintain. @DEEGAN2002 applied this framework to environmental and social disclosures, demonstrating how these practices reinforce an organization's legitimacy. Firms' legitimacy-seeking behaviors influence the depth and scope of disclosures, with firms in environmentally sensitive industries disclosing more to manage public perception.  

In the context of financial institutions, empirical studies show that ESG reporting is often used to satisfy stakeholder demands and enhance legitimacy. @THOMPSON2004 examines how banks respond to environmental stakeholders through reporting practices. Their findings reveal that bankers value annual reports for assessing corporate environmental impact, despite their limitations, and express some interest in expanded environmental disclosures. However, their focus primarily aims at safeguarding loans. Similarly, @SCHOLTENS2009 analyze how international banks balance different stakeholder interests through Corporate Social Responsibility (CSR) reporting, while @CARNEVALE2014 test the direct effects of the sustainability report on stock price in European markets, demonstrating that sustainability reports positively influence stock prices by signaling organizational compliance with evolving norms. @MURE2021 provide empirical evidence supporting stakeholder theory and legitimacy theory that ESG disclosure and performance can be used by banks as a tool to manage their reputation and maintain stakeholder trust, especially in the face of adverse events.


### Economic Theories

Economic theories, particularly information asymmetry and agency theory, explain how ESG reporting enhances market efficiency and reduces capital costs. @JENSEN1976 established the foundational agency theory framework, explaining the conflicts of interest between principals (shareholders) and agents (managers). This study lays the groundwork for understanding how ESG reporting can mitigate agency costs by aligning stakeholder interests. @AKERLOF1970 work on Information Asymmetry further explains how unequal information among market participants creates inefficiencies. @HART1987 provide a foundation for modern contract theory and influence the theory of the firm and organizational economics. This paper develops the principal-agent model, and demonstrate how disclosure requirements can act as screening devices. @GRAY1996 provide a theoretical framework for understanding the role of corporate social and environmental reporting in addressing information asymmetry and enhancing accountability. Applying these theories, studies have demonstrated that ESG disclosure reduces information gaps and lowers agency costs, thereby enhancing market efficiency. For instance, @DHALIWAL2011 and @ELGHOUL2011 find firms that initiate CSR disclosure and have superior CSR performance (relative to their industry peers) experience a subsequent reduction in their cost of equity capital, and attract more dedicated institutional investors and analyst coverage. Moreover, @CUI2018 observed that CSR engagement is inversely associated with reputation risk, largely due to reduced information asymmetry. 

### Institutional Isomorphism Framework

Institutional isomorphism framework explains the standardization and diffusion of reporting practices. @DIMAGGIO1983 identify three mechanisms driving institutional isomorphism: Coercive isomorphism, Mimetic isomorphism and Normative isomorphism. Their foundation study develops a theoretical framework explaining how these mechanisms operate within organizational fields - recognized areas of institutional life where organizations interact under similar pressures. This framework is valuable in understanding the standardization and diffusion of reporting practices across organizations. Coercive isomorphism explains how regulatory requirements and pressure from stakeholders drive adoption of standardized reporting formats. Mimetic isomorphism illuminates why organizations copy "best practice" reporting methods from industry leaders during periods of uncertainty. Normative isomorphism shows how professional networks and training lead to similar reporting approaches across organizations. Empirical research highlights the importance of institutional pressures, including coercive, normative, and mimetic forces, in driving firms to conform to ESG expectations [@DELMAS2004; @HIGGINS2014; @BEBBINGTON2018; @CHRISTENSEN2021]. 

### Emerging Theoretical Concepts

Emerging theoretical concepts and analytical lenses around natural capital and double materiality offer novel perspectives on integrating environmental considerations into financial decision-making. Natural capital concept, as articulated by @DASGUPTA2021 and @DAILY2009, emphasizes the valuation of biodiversity and natural capital, with implications for financial modeling [@ATKINSON2014]. @ECCLES2014 provide a theoretical foundation for integrated reporting,  which aligns closely with double materiality. The concept of "double-materiality" was first formally proposed by the European Commission [@EC2019] in Guidelines on Non-financial Reporting: Supplement on Reporting Climate-related Information published in June 2019; then was formalized by the regulation of the Corporate Sustainability Reporting Directive (CSRD)[see @EU_CSRD_2022] as a reporting requirement. It mandates a company to judge materiality from two perspectives 1) “the extent necessary for an understanding of the company’s development, performance and position” and “in the broad sense of affecting the value of the company”; 2) environmental and social impact of the company’s activities on a broad range of stakeholders. 

[^03]: Legitimacy is a generalized perception or assumption that the actions of an entity are desirable, proper, or appropriate within some socially constructed system of norms, values, beliefs, and definitions [@GINZEL2004; @NEILSEN1987; @PERROW1970].

ESG disclosure and reporting has a positive influence on a company's financial performance and risk. ESG reporting is not merely a compliance exercise but a strategic imperative for financial institutions.

## Empirical Research

### Firm Value and Performance

Stakeholder theories [see @FREEMAN1984; @DEEGAN2002] suggest that when organizations incorporate environmental, social, and governance (ESG) practices into their long-term strategic planning, they gain competitive advantages by addressing the needs of various groups - from employees and customers to government bodies and community members. This approach recognizes that business success is linked to creating value for all stakeholders, not just shareholders. The stakholders reward companys with good ESG practices through investment, consumption and higher productivity [@LI2018]In contrast, the trade-off hypothesis or traditionalist view [see @FRIEDMAN2007] suggests ESG practices negatively impact financial performance by increasing costs, harming corporate performance, and reducing competitive advantage. Scholars argue that focusing on social and environmental goals diverts managers from maximizing shareholder value. Similarly, satisfying non-shareholder stakeholders may hinder profit maximization and value creation for owners and managers[@GALANT2017].

Empirical studies on the relationship between ESG engagement and financial performance have provided conflicting findings. @BEN_LAHOUEL2019 and @BEN_LAHOUEL2021 find that firms with higher Corporate Social Performance (CSP) experience stronger financial performance, suggesting a mutually reinforcing relationship between social responsibility and profitability. Driven by stakeholder theory, @WU2013 find that ESG disclosures improve bank profitability by attracting ESG-conscious investors. Their findings attribute to stakeholder pressure for transparency. Drawing from empirical analysis using MSCI ESG STATS database ratings, @CORNETT2016 examine the relationship between the US commercial banks' corporate social responsibility (CSR) practices and financial performance. The study reveals that larger banks, despite facing criticism for profit-focused practices leading to the crisis, demonstrate consistently higher CSR strengths and concerns. Post-2009, these institutions showed marked improvement with increased CSR strengths and decreased concerns. Similarly, studies such as @CARNEVALE2014, @SHEN2016 and @BUALLAY2021 demonstrate banks' engagement in CSR activities can improve the financial performance. @TIAN2023 report that banks with green credit announcement show better market value and long-term performance. While @FERRERO2016 find a nonlinear relationship between ESG and financial performance employing listed companies in Europe. Moreover, @BUALLAY2023 investigate the effect of sustainability reporting on bank performance (operational, financial and market) in 7 regions that include 60 countries over the period 2008-2017. They find the negative relationship between ESG disclosures and banks' operational performance (ROA, ROE), and market performance (Tobin's Q). They argue that sustainability reporting may have negative impact on banks' asset utilization, in line with the trad-off theory [see @LEE2009].


### Bank Risk 

Despite growing scholarly attention to the link between CSR and bank performance, empirical research examining its influence on risk exposure within financial institutions remains limited. The existing literature primarily focuses on individual bank risk metrics. A study by @GANGI2019 examine 142 banks across 35 countries and find that stronger environmental engagement correlates with lower risk profiles as measured by Z-scores [see @LAEVEN2009]. Similarly, @SCHOLTENS2019 report a modest negative relationship between sustainability practices and standalone risk measures among European banks. Aligning with stakeholder theory, @DI_TOMMASO2020 demonstrate that higher ESG scores slightly curbed risk-taking among European banks (2007–2018), though this relationship was mediated by board composition factors including size, independent director presence, and gender diversity. @GANGWANI2024 report that higher ESG disclosure scores correlate with reduced commercial banking risks, specifically mitigating insolvency, leverage, and liquidity challenges. Further evidence from @CHIARAMONTE2022 demonstrate that both aggregate ESG scores and individual pillars enhanced bank stability during financial stress periods, with longer ESG disclosure histories amplifying these stabilizing effects. Expanding geographically, @NEITZERT2022 examine 582 international banks (2002–2018), corroborating the risk-mitigating effects of CSR activities on default and portfolio risks, consistent with risk management frameworks. However, their findings revealed heterogeneity across ESG pillars: environmental (E) initiatives exhibited the strongest and most consistent risk-reduction outcomes, whereas social (S) and governance (G) dimensions yielded less definitive results. Beyond the bank literature, @OIKONOMOU2012 highlights the risk-mitigating benefits of CSR, reporting that socially responsible firms show lower financial risk, while irresponsible behavior associated with increased risk.

Academic research on the relationship between systemic financial risk and ESG factors remains limited, with existing studies examining each of the three E, S and G pillars respectively. Regarding environmental (E) factors, @ESRB2016 points that an unanticipated rapid green transition could significantly impact asset values, with carbon-intensive investments facing declining profitability while low-carbon assets appreciate. In terms of governance (G) factors, @ANGINER2018 present compelling evidence that shareholder-friendly corporate governance structures correlate with increased bank systemic risk, as measured by Marginal Expected Shortfall (MES)[@ACHARYA2017] and SRISK metrics [@BROWNLEES2017].This relationship appears particularly pronounced for larger banks, potentially due to too-big-to-fail considerations, and is amplified in jurisdictions with more generous financial safety nets. Their findings derive from analyses of both U.S. banks (1990-2014) and international banks (2004-2008), suggesting the robustness of this relationship across different contexts and time periods.  

Using qualitative research method, @KUHN2022 investigates sustainable finance disclosures and initiatives in Germany. They find that ESG disclosures enable financial institutions to identify exposure to sectors with high dependency on vulnerable ecosystems. Disclosing these dependencies improves transparency and helps institutions avoid investments in activities contributing to ecological degradation such as deforestation-linked loans. Their study proposes integrating natural capital valuation into credit risk assessments to account for "hidden" risks. Case studies demonstrate that institutions adopting such frameworks reduce exposure to stranded assets and regulatory penalties.


### Investment Strategy and Portfolio Alignment

Banks have revised their approach to corporate social responsibility (CSR), placing greater emphasis on managing both direct and indirect risks associated with lending to firms facing environmental and social challenges [@CARNEVALE2012]. Drawn on the stakeholder theory, @GOSS2011 examine the link between CSR and the cost of bank loans. They find that firms with poor CSR performance face higher loan costs due to the creditor risk perceptions. 

@DEMETRIADES2025 investigate lending patterns between major banks and fossil fuel companies from 2001-2021 using global syndicated loan data. They find a complex dynamic where banks recognize and price in climate risks through higher interest rates and shorter loan terms, yet simultaneously increase loan volumes to brown firms. The findings suggest that regulatory pressure, particularly in Europe and the US, has led to more stringent lending policies, though not necessarily reduced lending volumes.

@BASU2022 report that high-ESG banks complement, rather than substitute, mortgage lending with community development investments in poor areas. However, banks are more likely to reject mortgage applications from these communities, suggesting social washing—using pro-social rhetoric while limiting actual support. Their findings align with the legitimacy theory, as banks appear to engage in CSR initiatives to maintain societal approval rather than genuinely addressing financial inclusion and social responsibility.


### Double Materiality

@ARAS2024 investigates the SDGs with ESG indicators through a double materiality perspective for 1888 companies from the OECD financial institutions. The study shows how commercial banks can identify and prioritize the SDGs and targets and how sustainable practices at the 
corporate level can contribute to achieving these global goals by adopting a sound approach. 



## Research Gap and Proposed Framework

In the early 2000s, @HOOGHIEMSTRA2000 argues that sustainability reporting research exhibits diverse and inconsistent findings, primarily due to the absence of a comprehensive theoretical framework. The research asserts that legitimacy was the dominant perspective. @SPENCE2010 identified stakeholder theory as the predominant and most effective framework for explaining sustainability reporting practices. They also point out that while many studies mention stakeholders in general, they do not explicitly reference stakeholder theory or other theoretical frameworks. Our review confirms their observations in the literature of banks' ESG disclosure. The majority of the studies show a preoccupation with stakeholder theory [@GALANT2017; @SHEN2016; @BUALLAY2021], legitimacy theory [e.g. @CARNEVALE2014], and to some extent also institutional theory [@HIGGINS2014; @BEBBINGTON2018; @CHRISTENSEN2021]. Moreover, these studies primarily rely on isolated theoretical frameworks rather than adopting a more holistic approach that integrates multiple theoretical perspectives on ESG disclosure. We propose an integrated Stakeholder-Legitimacy Framework of banks' ESG disclosure, which synthesizes and extends insights from @CAMPBELL2007 institutional-economic model and @AGUINIS2012 multilevel CSR analysis. At its core, the framework positions ESG disclosures as outcomes of dynamic negotiations between stakeholder pressures and legitimacy-seeking behaviors, moderated by banks' economic and institutional contexts. 


```{r stakeholder-legitimacy framework}
#| label: fig-flowchart
#| fig-cap: "ESG Disclosure Flowchart"

library(DiagrammeR)
library(DiagrammeRsvg)
library(rsvg)

grViz("
digraph {
  graph [
    rankdir = TB,
    size = '8,10!',        // Constrain graph dimensions
    ratio = 'compress',    // Compress to fit dimensions
    nodesep = 0.5,         // Reduce spacing between nodes
    ranksep = 0.4,         // Reduce vertical spacing
    fontsize = 10          // Reduce font size
  ]
  
  node [
    shape = oval,
    width = 1.5,          // Reduce node width
    height = 0.8,         // Reduce node height
    fontname = 'Helvetica'
  ]
  
  // Main vertical flow
  1 [label = 'Stakeholder Pressure']
  2 [label = 'Legitimacy-seeking Behavior']
  3 [label = 'ESG Disclosure']
  
  // Side nodes
  4 [label = 'Bank Economic Context']
  5 [label = 'Bank Institutional Context']
  
  // Main vertical connections
  1 -> 2 -> 3 [minlen = 1]
  
  // Side connections
  4 -> 3 [constraint = false]
  5 -> 3 [constraint = false]
  
  // Force side nodes to same rank
  {rank = same; 4; 5}
  
  // Position side nodes using invisible edges
  invisible_node [shape = point, width = 0, height = 0]
  2 -> invisible_node [style = invis]
  invisible_node -> 4 [style = invis]
  invisible_node -> 5 [style = invis]
}
") %>% 
  export_svg() %>% 
  charToRaw() %>% 
  rsvg_pdf("flowchart.pdf")

knitr::include_graphics("flowchart.pdf")

```


In the integrated framework (@fig-flowchart), stakeholder pressures act as catalysts for banks' ESG disclosures. According to stakeholder theory, companies should consider and balance the diverse viewpoints and expectations of all groups affected by or invested in their operations, rather than focusing solely on shareholders [@BUCHHOLZ2005; @LAPLUME2008]. @FREEMAN1984 suggests that company management should stay attuned to changing dynamics and trends affecting both their organization's internal constituents and outside parties. Stakeholder pressures arise from key external actors, including regulators, investors, and customers. These actors demand greater transparency and accountability, motivating banks to provide valuable explanations of how they answer to the societal call for sustainable business conduct. Banks respond to these pressures through strategic legitimacy-seeking behaviors. Banks' ESG disclosures range from symbolic gestures (e.g. adopting TCFD guidelines without operational changes) to substantive actions such as phasing out coal financing. Crucially, legitimacy is not a static achievement but an ongoing process; banks must continuously adapt to shifting norms, such as the rise of the TNFD as a biodiversity benchmark. In the stakeholder-legitimacy context, influential stakeholders execute greater pressure on companies to explain and justify their business conduct. Therefore, sustainable reporting and disclosed ESG information serve as a way for companies to establish and maintain their legitimacy in the eyes of these stakeholders [@CAMPBELL_D2003]. 

The integrated framework further incorporates moderating contexts that shape ESG disclosures produced by the translation of stakeholder pressures into banks' legitimacy-seeking behaviors, including economic conditions such as size and profitability, and institutional context such as ownership structure and regulatory regimes. 


### Bank Size and Financial Perfromance

Firm size (usually measured by total assets or market capitalization) can be considered to have a positive effect on the adoption and extent of ESG disclosure. As ESG engagements involve complex processes and require a large scale to be effective [@YOUN2015], firm size is a critical factor in making them successful. Furthermore, according to academic research, company visibility plays a crucial role in engaging a broader and more heterogeneous set of stakeholders [@PELOZA2006]. Visibility encourages organizations to adhere to social pressures and to be more sensitive to stakeholders' expectations. And firm size is positively associated with company visability [@MEZNAR1995]. Therefore, companies with significant market visibility tend to prioritize enhanced sustainability initiatives, motivated by the desire to maintain a positive reputation among their stakeholders [@DAMATO2020; @ABDI2022]. The empirical studies of @FAVINO2019 and @ALAREENI2020 support this viewpoint. They conclude that firm size is an important determinant of ESG scores, as larger firms have more available resources to provide ESG data and are more likely to have their ESG data available. @DREMPETIC2020 find that larger firms tend to have higher ESG scores compared to smaller firms. They also argue that size bias in ESG ratings practice, with larger firms receiving systematically better ESG evaluations. @BUALLAY2019 and @LAMANDA2024 find consistent empirical results in banking sector. 

A body of literature study the relation between ESG disclosure/performance and financial performance, mostly focusing on the impact of ESG activities on financial performance [@FAVINO2019; @BUALLAY2019; @DAMATO2020; @ABDI2022; @LAMANDA2024]. Financial performance as a determinant of ESG disclosure receives less academic attention. Due to their higher ability and flexibility of handling the costs of disclosing, companies with higher profitability (lower leverage and gearing) are often presumed by researchers to be better positioned to cope with any negative impacts that might arise from revealing unfavorable information in their ESG disclosure. However, the empirical results are mixed. @BRAMMER2008 report that firms with higher leverage are less likely to make voluntary environmental disclosures due to financial constraints and costs. The positive relationship between profitability (measured by return on assets (ROA), or return on equity (ROE)) are reported by @HADDOCK2005, @LIU2009 and @GAMERSCHLAG2011. Their studies argue that profitable companies are more likely to disclose ESG information as they have more resources available to absorb the associated costs. While @DYDUCH2017 do not find a significant relationship between financial performance and sustainability reporting in Polish listed companies. @YUEN2022 report a U-shaped relationship between ESG activities and bank profitability.   


### Ownership and Regulation  

Ownership structure and regulation can be considered as external influential factors that are beyond management control. Studies examining different ownership variables have emerged primarily in recent years. The most investigated ownership structure include a company's listing status (publicly listed), institutional shareholding, government involvement (state-owned), and shareholding concentration. 

Publicly listed companies, in order to comply with regulations and to be aligned with industry best practices, as well as to stay attuned to stakeholder pressure and see societal legitimacy, tend to be actively engaged in ESG reporting. @HADDOCK2005 reports that public offering is linked to greater adoption of social reporting practices. Additionally, publicly listed companies tend to disclose more sustainability-related information [@GAMERSCHLAG2011]. The publicly listed ownership structure was primarily examined as a variable in studies conducted before 2015. However, around 2020, publicly listed companies were more commonly used as data samples rather than as a subject of ownership structure analysis [e.g. @SAHASRANAMAM2020; @AMEEN2022; @ALOBAID2024]. In more recent research, institutional shareholding has emerged as a key variable of interest, such as @DYCK2019, @CHEN2020, and @AMEEN2022. Institutional investors are defined as pension funds, insurance companies, banks, sovereign wealth funds, and other institutions that manage and invest funds on behalf of their beneficiaries; they are major players in the global financial markets [@AMEEN2022]. Aligned with agency theory, @SHLEIFER1986 posit that institutional investors, leveraging their significant equity stakes and specialized expertise, closely monitor corporate operations. This oversight helps mitigate agency costs by ensuring management decisions align with shareholder priorities, ultimately promoting more efficient investment practices such as facilitating takeovers. Due to their influential position, institutional investors can shape companies' ESG initiatives by execute the stakeholder pressure on management to disclose greater ESG-related data, aiming to fulfill non-financial goals and bolster both their own and the firms' public image [@DYCK2019; @GARCIA-SANCHEZ2020]. Nevertheless, empirical literature provides inconsistent findings. Some studies indicate that there is a positive association between institutional ownership and ESG performance [@DYCK2019; @CHEN2020; @AMEEN2022; @RAIMO2020]. @ZHOU2019 repots that higher institutional ownership can facilitate firms’ transparency by providing voluntary CSR reports. @ALUCHNA2022 find a negative association between institutional ownership and disclosure of the social performance.  @QU2007 produces insignificant results on the relationship between corporate ownership, including institutional ownership and state ownership, and a company’s CSR engagement. 

While institutional investors’ role in ESG practices has garnered significant attention, the influence of state ownership introduces distinct dynamics, particularly in balancing public policy objectives with corporate accountability and transparency. @TAGESSON2009 find that State-owned corporations disclose more social information than privately owned corporations. One explanation is that state-owned corporations face heightened scrutiny, with pressure from their government stakeholders and public/media attention driving demands for accountability. These entities appear to have responded by aligning with stakeholder demands, reflecting compliance with external expectations. This fiding is also aligned with the social view of state ownership theory [see @STIGLITZ1993] that state ownership addresses market failures and improve social welfare.  @LAU2016 and @ZHOU2019 support the social view that state ownership has a significant positive impact on firms’ CSR disclosure decisions. @BOSE2018 demonstrate that higher institutional ownership leads to higher green banking disclosure. However, @RAIMO2020 report a negative relationship between state ownership and integrated report quality. Shareholding concentration can mean that one or several shareholders have substantial minority ownership stakes and voting rights, such as 10 or 20 percent [@SHLEIFER1997]. Corporate governance literature highlights the frequent comprimise of minority shareholder rights in firms dominant controlling shareholders, where concentrated ownership can lead to power imbalances and disproportionate influence over strategic decisions [@SMITH2022]. Therefore, @SMITH2022 demonstrate that controlling shareholders are less motivated to provide CSR information, with minority shareholders have less power in CSR disclosure practices. @BORGES2024 report that large sharehoders have low interest in the firm's CSR reporting. This result corroborates the finding by @DUCASSY2015 that dominant shareholders lack incentives to allocate resources to social initiatives in contexts with limited stakeholder pressures to legitimize corporate operations, which aligns with stakeholder theory and legitimacy theory. @LAU2016 also report that higher ownership concentration negatively affects CSR performance.  

Studies that exmaine the influence of ESG regulations are still scarce. @HESS2007 suggests that current vulantary initialtives of social reporting are insufficienct to achieve coporate accountability. The study argues, aligned with ligitamacy theory, that social reporting is often driven by strategic considerations rather than a genuine commitment to accountability. Companies may engage in social reporting as a form of "impression management" to portray a positive image, rather than to provide meaningful and transparent information to stakeholders. @HESS_etal2007 suggest that an overarching mendatory social reporting system as the appropriate reporting system to avoid corporate strategic disclosing behavior. @QU2007 argues that the robustness and enforcement of government regulations critically shape the scope and depth of corporate social responsibility (CSR) initiatives, underscoring regulatory frameworks as a catalyst in advancing CSR practices by institutionalizing corporate accountability and alignment with societal expectations. @DESAI2024 reports that stock market investors reacts positively to the mandatory ESG disclosure regulation because mandatory ESG disclosure is considered as a positive signal by reducing information asymmetry. @KRUEGER2024 demonstrate that mandatroy ESG disclosure improve firms' liquidity, mitigating adverse selection problem. They also observe improvements in liquidity as markets anticipate the introduction of mandatory ESG disclosure regulations, reflecting proactive adjustments to expected regulatory changes. @LEE2024 find that compulsory ESG disclosure regulation signifies the negative impact of ESG scores on firms' extreme risks; the effect is amplified in cases where firms issue green bonds. @CICCHIELLO2023 examine the effects of NFRD. They argue that government-mandated ESG disclosure regulations targeting corporate social and environmental impacts are proven to strengthen both the rigor and reliability of corporate disclosure practices, fostering greater societal accountability. @BOSE2018 demonstate that green banking regulatory framework has positive association with green banking disclosure practices. However, @MANOS2024 report that self-regulated banks have higher ESG scores by adopting Principles for Responsible Banking (PRB). Additionally, their findings reveal the negative impact of regulatory quality on ESG disclosure performance. They argue that banks use mandatory ESG disclosure as a leverage to seek ligitimacy in weaker institutional environment. 

This section presents an integrated Stakeholder-Legitimacy Framework, tailored to the banking sector, to explain banks’ ESG disclosure practices, synthesizing stakeholder theory, legitimacy theory, and institutional contexts into a cohesive analytical pathway. The framework posits that ESG disclosures arise from dynamic interactions between external stakeholder pressures (e.g., regulators, investors, customers) and banks’ legitimacy-seeking behaviors, moderated by economic factors (e.g., size, profitability) and institutional conditions (e.g., ownership structure, regulatory regimes). By bridging insights from @CAMPBELL2007 and @AGUINIS2012, the framework advances a holistic view of ESG reporting as a negotiated outcome of strategic adaptation to stakeholder demands and institutional expectations.

Critical research gaps persist, however. First, while institutional and state ownership structures are increasingly scrutinized, findings remain inconsistent: institutional ownership’s role in ESG performance varies across studies [e.g., @DYCK2019; @ALUCHNA2022], and state ownership’s dual mandate for social welfare and corporate accountability lacks consensus [@TAGESSON2009; @RAIMO2020]. Second, the moderating role of regulatory frameworks is underexplored, particularly in the highly regulated banking sector. Empirical research on how mandatory ESG regulations (e.g., CSRD, EU Taxonomy) shape bank risk profiles, financial performance, and governance mechanisms remains scarce, despite the systemic importance of banks and their unique exposure to ESG-related risks. While studies have begun analyzing disclosure quality versus symbolic compliance in non-financial sectors [@CICCHIELLO2023; @MANOS2024], sector-specific analyses for banking—where regulatory pressures intersect with complex risk intermediation roles—are urgently needed. Third, compounding this complexity, banks’ dual role in ESG ecosystems warrants deeper scrutiny: as both reporting entities subject to disclosure mandates and financing intermediaries influencing firms’ access to green capital, they create transmission effects that amplify regulatory impacts across economies. For instance, @WANG2023 demonstrates that ESG disclosure regulations propagate through bank lending networks, altering borrowers’ environmental practices even without direct regulatory coverage—a phenomenon underexplored in the literature. Fourth, mixed evidence on the relationship between financial performance and ESG disclosure—such as conflicting results on profitability [@HADDOCK2005; @DYDUCH2017]—highlights the need for contextualized analyses. Finally, concentrated ownership’s tendency to marginalize minority shareholder influence on CSR practices [@SMITH2022; @BORGES2024] underscores a governance gap in aligning dominant shareholders with broader stakeholder expectations. These gaps signal opportunities for future research to refine the framework, particularly in disentangling ownership dynamics, regulatory efficacy, and the interplay between symbolic and substantive ESG strategies across diverse institutional settings.