---
title: "A Stakeholder-Legitimacy Framework for ESG Disclosure in Banking Under Evolving Regulatory Mandates"
date: today
dateformat: long
author: 
  - name: Ying (Veronica) Zhang
    email: veronica.zhang@qub.ac.uk
    affiliations: 
      - id: sbs
        name: Queen's University Belfast
        department: Institute for Global Food Security (IGFS) School of Biological Sciences
        address: Belfast, UK
    attributes:
      corresponding: true
    note: |
      This work was supported by the BIOFIN-EU Project (Grant agreement 101135476, UKRI through the UK Horizon Guarantee Fund (no.10092956) and through the National Environment Research Council (Grant no. Z503368)  
  - name: Lisa Sheenan
    affiliations: 
      - id: ucd
        name: University College Dublin
        department: UCD Michael Smurfit Graduate Business School
        address: Dublin, Ireland
      - ref: sbs
  - name: Fearghal Kearney
    affiliations: 
      - id: qbs
        name: Queen's University Belfast
        department: Queen's Business School
        address: Belfast, UK
  - name: Junyu Zhang
    affiliations: 
      - ref: sbs  
  - name: George Hutchinson
    affiliations: 
      - ref: sbs
  - name: Neil Reid
    affiliations: 
      - ref: sbs
      - ref: qbs
      - id: cbw
        name: Queen's University Belfast
        department: Co-Centre for Climate + Biodiversity + Water, School of Biological Sciences
        address: Belfast, UK
  - name: Paul Brereton
    affiliations: 
      - ref: sbs          

abstract: |
  The global financial sector is undergoing a transformative shift as ​material risks stemming from environmental and social challenges, such as​ biodiversity loss, ecosystem degradation, ​and climate change,​​ underscore the growing importance of Environmental, Social, and Governance (ESG) factors for economic stability. By reviewing the literature, this study advances a novel Stakeholder-Legitimacy Framework, integrating stakeholder theory, legitimacy theory, and institutional isomorphism to explain how banks’ ESG disclosures emerge from dynamic negotiations between external pressures (e.g., regulators, investors) and strategic legitimacy-seeking behaviours. The study reveals that while ESG disclosures mitigate bank-specific risks, conflicting evidence persists on their financial performance impacts. Critical gaps include underexplored transmission effects of regulations through lending networks, inconsistent roles of institutional and state ownership, and tensions between symbolic compliance and substantive action.
keywords: 
  - ESG disclosure
  - IFRS S1 and S2
  - Stakeholder-Legitimacy framework
  - bank risk
  - ownership
  - climate change
  - JEL code M14, M41, G28
bibliography: review.bib
format: 
  elsevier-pdf: 
    keep-tex: true
    header-includes: |
      \usepackage{graphicx}
      \usepackage{adjustbox}
      \usepackage{float}
      \floatplacement{figure}{H}
    journal: 
      name: "The British Accounting Review"
      cite-style: authoryear
    geometry: [left=3.5cm, right=3.5cm, top=4.5cm, bottom=4.5cm]
execute: 
  eval: true
  echo: false
  output: true
  warning: false
  error: false
  include: true
---

```{r set up}
#| label: libraries
#| output: false
#| include: false

library(tidyverse)
library(kableExtra)
library(DiagrammeR)
library(DiagrammeRsvg)
library(rsvg)
library(knitr)
library(tinytex)




```

\newpage

# Introduction

The global financial system is undergoing a seismic shift as nature-related risks, ranging from biodiversity loss to ecosystem degradation, emerge as material threats to economic stability [@UNEP2023]. Over the past decade, voluntary sustainability frameworks like the Task Force on Climate-related Financial Disclosures (TCFD) have evolved into robust, jurisdictionally mandated systems, driven by regulatory imperatives to combat greenwashing, harmonise reporting standards, and align financial flows with planetary boundaries. Central to this transformation are frameworks such as the Task Force on Nature-Related Financial Disclosures (TNFD), the International Sustainability Standards Board (ISSB), and region-specific regulations like the EU’s Corporate Sustainability Reporting Directive (CSRD) and the UK’s Sustainability Disclosure Requirements (SDR) (See the appendix table). These frameworks reflect a growing consensus that environmental risks are inextricably linked to financial performance, necessitating disclosures that integrate double materiality, assessing both financial risks to firms and firms’ impacts on ecosystems.

Banks are uniquely exposed to systemic ESG risks due to their role as financial intermediaries, warranting specialized regulatory analysis. Nevertheless, empirical research on how ESG rules influence their risk exposure, governance, and performance remains fragmented. Critically, this fragmentation obscures how current ESG disclosures often become performative tools rather than transparency mechanisms, masking actual environmental exposures through selective reporting or "sustainability-washing" [@SMITH2022]. Existing studies disproportionately focus on non-financial firms [e.g. @YOUN2015; @FAVINO2019; @BUALLAY2019; @DYCK2019], leaving critical gaps in understanding banks’ unique intermediation role. For instance, while the TNFD’s LEAP methodology and the ISSB’s IFRS S1 and S2 standards are gaining traction, their sector-specific implications for banking, such as transmission effects through lending networks or the tension between symbolic compliance and substantive action, are often misunderstood or inadequately captured in disclosures, leading to misleading risk assessments.​​ Theoretical frameworks also remain fragmented, with stakeholder theory, legitimacy theory, and institutional isomorphism often treated in isolation rather than as interconnected drivers of ESG disclosure.[^01] This theoretical disconnection perpetuates misleading siloed analyses of bank ESG behaviors, failing to reflect how these forces dynamically intersect.​ This paper addresses these gaps by synthesising empirical evidence, proposing an integrated theoretical model, and identifying urgent research priorities at the nexus of finance, regulation, and ecological stewardship.

[^01]: Stakeholder theory [@FREEMAN1984; @MITCHELL1997], legitimacy theory [@SUCHMAN1995], and institutional isomorphism [@DIMAGGIO1983] have been applied individually to explain ESG disclosure motivations. Yet, to our knowledge, ​no major study​ has synthesised these three frameworks to examine their synergetic effects, creating a theoretical silo problem. This fragmentation mirrors broader critiques of theoretical dis-integration in ESG disclosure [@ECCLES2012].

Current empirical studies show how banks use ESG reporting to navigate real-world constraints. @THOMPSON2004 examines how banks respond to environmental stakeholders through reporting practices. Their findings reveal that bankers value annual reports for assessing corporate environmental impact, despite their limitations, and express some interest in expanded environmental disclosures. However, their focus primarily aims at safeguarding loans, highlighting how disclosures are often deployed as reputational shields rather than accountability tools.​​ Their study reflects tensions between stakeholder demands and bank risk management. Similarly, @SCHOLTENS2009 analyse how international banks balance different stakeholder interests through Corporate Social Responsibility (CSR) reporting, while @CARNEVALE2014 test the direct effects of the sustainability report on stock price in European markets, demonstrating that sustainability reports positively influence stock prices by signaling organizational compliance with evolving norms. @MURE2021 provide empirical evidence supporting stakeholder theory and legitimacy theory that ESG disclosure and performance can be used by banks as a tool to manage their reputation and maintain stakeholder trust, especially in the face of adverse events. Collectively, these studies expose a fundamental problem: ESG disclosures are frequently decoupled from substantive action, engineered to signal legitimacy to stakeholders while masking underlying tensions.​​ These findings underscore a critical gap: existing theories fail to fully explain how banks reconcile competing stakeholder, legitimacy, and economic imperatives in ESG implementation. With mandatory frameworks expanding globally, this disconnect necessitates an integrated conceptual lens now to decode the strategic interplay between disclosure motivations and outcomes.​​ This necessitates the integrated Stakeholder-Legitimacy Framework proposed in Section IV.  

In this study, we advance a novel Stakeholder-Legitimacy Framework [@CAMPBELL2007; @AGUINIS2012] tailored to banks, synthesising stakeholder theory, legitimacy theory, and institutional isomorphism. This model positions ESG disclosures as dynamic outcomes of external pressures (e.g., regulators, shareholders) and banks’ legitimacy-seeking behaviors, moderated by economic factors (e.g., bank size) and institutional contexts (e.g., ownership structures). Our synthesis of extant literature highlights research gaps regarding ownership dynamics, regulatory efficacy, and systemic risk implications. The proposed Stakeholder-Legitimacy Framework offers a holistic lens to analyse banks’ ESG disclosures as context-dependent processes shaped by stakeholder pressures and legitimacy imperatives. 

Our study advances sustainable finance research by contributing significantly across three interconnected domains.​​ First, we introduce theoretical innovation through the Stakeholder-Legitimacy Framework, which bridges siloed theoretical perspectives [@ECCLES2012] to holistically conceptualise banks’ ESG disclosures as dynamic, context-dependent processes. This framework reframes disclosures as outcomes shaped by external stakeholder pressures (e.g., regulators, activist investors) and internal legitimacy-seeking behaviors, moderated by institutional factors like ownership structures. Second, we offer actionable policy relevance by proposing concrete steps for regulators and policymakers, such as integrating natural capital valuation into stress testing and harmonizing ISSB-TNFD reporting interoperability. These recommendations address concerns about symbolic compliance [@SMITH2022] and aim to align regulatory frameworks with planetary boundary imperatives. Third, we identify underexplored dimensions of banking-sector ESG research including empirical inconsistencies regarding ownership dynamics and the efficacy of mandatory frameworks (e.g., CSRD, EU Taxonomy) in shaping bank risk profiles and governance [@ESRB2016]. 

The remainder of the literature review is organized as follows: In Section II and III, we review the theories underpinning, and the empirical studies on the relationship between ESG disclosure and the banking sector, respectively. In Section IV, we propose the integrated Stakeholder-Legitimacy framework applied to the banking industry and review the research gaps. Section V is the conclusion section. An overview of the nature-related financial disclosure frameworks from different dimensions, as background information of this review, is provided in the appendix. 


# Theoretical Foundation

The relationship between environmental, social, and governance (ESG) reporting and financial institutions has garnered increasing scholarly attention, driven by global sustainability challenges. This section first establishes the conceptual landscape of contemporary ESG disclosure before synthesising theoretical frameworks that explain how and why ESG reporting influences financial institutions’ operations, decision-making, and market outcomes. 


## Conceptual Advances in ESG Disclosure

Contemporary theoretical concepts and analytical lenses around natural capital and double materiality offer novel perspectives on integrating environmental considerations into financial decision-making. ESG reporting is not merely a compliance exercise but a strategic imperative for financial institutions. Natural capital concept, as articulated by @DASGUPTA2021, emphasises the valuation of biodiversity and natural capital, with implications for financial modelling [@ATKINSON2014]. @ECCLES2014 provide a theoretical foundation for integrated reporting, which aligns closely with double materiality. The concept of "double-materiality" was first formally proposed by the European Commission [@EC2019] in Guidelines on Non-financial Reporting: Supplement on Reporting Climate-related Information published in June 2019; then was formalised by the regulation of the Corporate Sustainability Reporting Directive (CSRD)[see @EU_CSRD_2022] as a reporting requirement. It mandates a company to judge materiality from two perspectives 1) “the extent necessary for an understanding of the company’s development, performance and position” and “in the broad sense of affecting the value of the company”; 2) environmental and social impact of the company’s activities on a broad range of stakeholders. 


## Organisational Theories

Organisational theories, particularly stakeholder theory and legitimacy theory, highlight the institutional motivations and external pressures driving ESG reporting practices.  Stakeholder theory, developed by @FREEMAN1984, posits that organisations are accountable to a broad range of stakeholders beyond shareholders, including regulators, customers (investors), employees, and civil society. Building on this, @MITCHELL1997 proposed the theory of stakeholder salience emphasising understanding why and how managers prioritise certain stakeholders, rather than prescribing whom they should prioritise. Their theory provides tools to identify stakeholder types via a typology and explain managerial responses through the concept of salience, which assesses stakeholders' legitimacy, power, and urgency. This typology offers a framework for financial institutions to navigate competing stakeholder interests effectively. @DONALDSON1995 provided theoretical justification for stakeholder theory, emphasising its descriptive, instrumental, and normative validity in explaining corporate behaviour. 

Legitimacy theory, foundationally developed by @SUCHMAN1995, identifies three primary forms of legitimacy[^03], pragmatic, moral, and cognitive, that organisations seek to maintain. @DEEGAN2002 applied this framework to environmental and social disclosures, demonstrating how these practices reinforce an organisation's legitimacy. Firms' legitimacy-seeking behaviours influence the depth and scope of disclosures, with firms in environmentally sensitive industries disclosing more to manage public perception. In banking contexts, ESG disclosures function as legitimacy-management tools to address stakeholder pressures, as evidenced by sector-specific studies [see @THOMPSON2004; @MURE2021; @CARNEVALE2014].  

Building on stakeholder theory and legitimacy theory, institutional isomorphism elucidates the structured mechanisms that drive organisational conformity. @DIMAGGIO1983 identify three isomorphic drivers that create organisational homogeneity:

1. **Coercive isomorphism**: explains how regulatory requirements and pressure from stakeholders drive adoption of standardised reporting formats. 

2. **Mimetic isomorphism**: illuminates why organisations copy "best practice" reporting methods from industry leaders during periods of uncertainty. 

3. **Normative isomorphism**: shows how professional networks and training lead to similar reporting approaches across organisations.

Their foundation study develops a theoretical framework explaining how these mechanisms operate within organisational fields. This framework is valuable in understanding the standardisation and diffusion of reporting practices across organisations. Empirical research highlights the importance of institutional pressures, including coercive, normative, and mimetic forces, in driving firms to conform to ESG expectations. @DELMAS2004 demonstrate firms' heterogeneous interpretations of isomorphic forces determine substantive versus symbolic ESG integration. This divergence mirrors banking practices where identical EU CSRD mandates yield divergent TNFD implementation depth. @BEBBINGTON2018 position normative isomorphism as instrumental in bridging accounting praxis with planetary boundaries, arguing that professional networks institutionalise "double materiality" reporting, which notes the field's underdevelopment in translating SDGs into bank capital allocation. @CHRISTENSEN2021 reveal mimetic isomorphism's unintended consequences: firms under US/EU disclosure mandates increasingly adopt boilerplate ESG language to signal conformity while obscuring institution-specific ecological impacts.  


## Economic Theories

Economic theories, particularly information asymmetry and agency theory, explain how ESG reporting enhances market efficiency and reduces capital costs. @JENSEN1976 established the foundational agency theory framework, explaining the conflicts of interest between principals (shareholders) and agents (managers). This study lays the groundwork for understanding how ESG reporting can mitigate agency costs by aligning stakeholder interests. @AKERLOF1970 work on Information Asymmetry further explains how unequal information among market participants creates inefficiencies. @HART1987 provide a foundation for modern contract theory and influence the theory of the firm and organisational economics. This paper develops the principal-agent model, and demonstrate how disclosure requirements can act as screening devices. @GRAY1996 provide a theoretical framework for understanding the role of corporate social and environmental reporting in addressing information asymmetry and enhancing accountability.

[^03]: Legitimacy is a generalised perception or assumption that the actions of an entity are desirable, proper, or appropriate within some socially constructed system of norms, values, beliefs, and definitions [@GINZEL2000; @NEILSEN1987; @PERROW1970].


##  Integrating Theoretical Frameworks: Addressing Complexity in Banking

While each theoretical lens offers valuable insights, the complex and often contradictory nature of ESG impacts on financial institutions necessitates an integrated framework. Relying solely on one perspective provides an incomplete picture. For example, Stakeholder theory explains why banks respond to pressures but struggles to predict how (substantive vs. symbolic). Legitimacy theory explains conformity motives but less so the strategic heterogeneity observed. Economic theories focus on market outcomes but may underplay the role of institutional pressures and legitimacy management unique to highly regulated, systemically important banks. Furthermore, banks operate under intense regulatory scrutiny, manage complex information asymmetries (e.g., in lending), face diverse and powerful stakeholders (regulators, depositors, investors), and are central to allocating capital towards sustainability goals. This confluence of factors creates unique tensions. For instance, banks may need to balance short-term shareholder returns with long-term systemic risks related to climate change or social inequality.

Therefore, synthesising organisational and economic theories offers a more robust, multi-dimensional lens. It captures the interplay between external pressures internal strategic choices, and resulting market consequences. This integrated approach is essential for understanding the nuanced and often paradoxical ways ESG reporting shapes the modern banking landscape.


# Extant Empirical Evidence

## Firm Value and Performance

Stakeholder theories suggest that when organisations incorporate environmental, social, and governance (ESG) practices into their long-term strategic planning, they gain competitive advantages by addressing the needs of various groups - from employees and customers to government bodies and community members. This approach recognises that business success is linked to creating value for all stakeholders, not just shareholders. The stakeholders reward companies with good ESG practices through investment, consumption and higher productivity [@LI2018]. In contrast, the trade-off hypothesis or traditionalist view [see @FRIEDMAN2007] suggests ESG practices negatively impact financial performance by increasing costs, harming corporate performance, and reducing competitive advantage. Scholars argue that focusing on social and environmental goals diverts managers from maximising shareholder value. Similarly, satisfying non-shareholder stakeholders may hinder profit maximization and value creation for owners and managers[@GALANT2017].

Empirical studies on the relationship between ESG engagement and financial performance have provided conflicting findings. Driven by stakeholder theory, @WU2013 find that ESG disclosures improve bank profitability by attracting ESG-conscious investors. Their findings attribute to stakeholder pressure for transparency. Drawing from empirical analysis using MSCI ESG STATS database ratings, @CORNETT2016 examine the relationship between the US commercial banks' corporate social responsibility (CSR) practices and financial performance. The study reveals that larger banks, despite facing criticism for profit-focused practices leading to the crisis, demonstrate consistently higher CSR strengths and concerns. Post-2009, these institutions showed marked improvement with increased CSR strengths and decreased concerns. Similarly, studies such as @CARNEVALE2014, @SHEN2016 and @BUALLAY2021 demonstrate banks' engagement in CSR activities can improve the financial performance. @TIAN2023 report that banks with green credit announcement show better market value and long-term performance. While @FERRERO2016 find a nonlinear relationship between ESG and financial performance employing listed companies in Europe. Moreover, @BUALLAY2023 investigate the effect of sustainability reporting on bank performance (operational, financial and market) in 7 regions that include 60 countries over the period 2008-2017. They find the negative relationship between ESG disclosures and banks' operational performance (ROA, ROE), and market performance (Tobin's Q). They argue that sustainability reporting may have negative impact on banks' asset utilisation, in line with the trad-off theory [see @LEE2009]. @ARAS2024 investigates the SDGs with ESG indicators through a double materiality perspective for 1888 companies from the OECD financial institutions. The study shows how commercial banks can identify and prioritise the SDGs and targets and how sustainable practices at the corporate level can contribute to achieving these global goals by adopting a sound approach.


## Bank Risk 

Despite growing scholarly attention to the link between CSR and bank performance, empirical research examining its influence on risk exposure within financial institutions remains limited. The existing literature primarily focuses on individual bank risk metrics. Research generally indicates a negative relationship between stronger ESG performance/disclosure and individual bank risk. Studies across various regions and time periods have found associations with lower insolvency risk (Z-scores), reduced leverage and liquidity challenges, and enhanced stability during financial stress. The length of a bank's ESG reporting history appears to amplify these stabilizing effects. However, findings suggest heterogeneity across ESG pillars, with environmental (E) factors often demonstrating the strongest and most consistent risk-mitigation outcomes compared to social (S) and governance (G) dimensions.

Academic research on the relationship between systemic financial risk and ESG factors remains limited, with existing studies examining each of the three E, S and G pillars respectively. Research highlights potential vulnerabilities, such as the risk of asset value destabilization from an unanticipated rapid green transition affecting carbon-intensive sectors. Conversely, studies focusing on governance factors present a concerning counterpoint: shareholder-friendly governance structures have been associated with increased systemic risk, particularly for larger banks, potentially exacerbated by too-big-to-fail perceptions and generous safety nets. Qualitative research also offers insights, suggesting that ESG disclosures, particularly those integrating natural capital considerations, can help banks identify hidden ecological dependencies and reduce exposure to stranded assets and regulatory penalties in sectors like deforestation-linked finance.

The key empirical findings linking ESG disclosure to both individual and systemic bank risk are summarised in @tbl-literature.

```{r literature table}
#| label: tbl-literature
#| tbl-cap: "Summary of Empirical Evidence Linking ESG Disclosure to Individual and Systemic Bank Risk"

df <- tibble(
  "Risk Type" = c("Individual Bank Risk", 
                  "Individual Bank Risk", 
                  "Individual Bank Risk", 
                  "Systemic Financial Risk", 
                  "Systemic Financial Risk", 
                  "Qualitative Risk Insights"
                  ), 
  "Findings" = c("Higher ESG scores linked to lower insolvency, leverage, and liquidity risk (e.g Gangi et al., 2019)",
                  "Longer ESG reporting history enhances risk mitigation (Chiaramonte et al., 2022)",
                  "Environmental (E) pillar shows strongest risk-reduction effect (Neizert and Petras, 2022)", 
                  "Rapid green transition can destabilise assets (ESRB, 2016)", 
                  "Shareholder-friendly governance may increase systemic risk (Anginer et al., 2018)", 
                  "Natural capital integration may reduce exposure to ecological risks (Kuhn, 2022)"
                  ), 
  "Implication" = c("Supports ESG as a risk management tool", 
                    "Suggests consistency and time horizon of disclosure matters", 
                    "Emphasis importance of environmental reporting", 
                    "Calls for climate transition scenario analysis", 
                    "Raises concerns for governance-driven risk in large banks", 
                    "Points to improved credit risk assessment via ESG reporting"
                  ) 
)


df_kbl <- kbl(df, 
              booktabs = TRUE, 
              format = "latex") |>
          kable_styling(latex_options = "striped", 
                        font_size = 9) |>
          row_spec(0, bold = TRUE) |>
          column_spec(1, width = "2cm") |>
          column_spec(2, width = "5cm") |>
          column_spec(3, width = "5cm") 

df_kbl

```



## Bank Lending, ESG Risk and Strategic Alignment

Banks have revised their approach to corporate social responsibility (CSR), placing greater emphasis on managing both direct and indirect risks associated with lending to firms facing environmental and social challenges [@CARNEVALE2012]. Drawn on the stakeholder theory, @GOSS2011 examine the link between CSR and the cost of bank loans. They find that firms with poor CSR performance face higher loan costs due to the creditor risk perceptions. 

@DEMETRIADES2025 investigate lending patterns between major banks and fossil fuel companies from 2001-2021 using global syndicated loan data. They find a complex dynamic where banks recognize and price in climate risks through higher interest rates and shorter loan terms, yet simultaneously increase loan volumes to brown firms. The findings suggest that regulatory pressure, particularly in Europe and the US, has led to more stringent lending policies, though not necessarily reduced lending volumes.

@BASU2022 report that high-ESG banks complement, rather than substitute, mortgage lending with community development investments in poor areas. However, banks are more likely to reject mortgage applications from these communities, suggesting social washing—using pro-social rhetoric while limiting actual support. Their findings align with the legitimacy theory, as banks appear to engage in CSR initiatives to maintain societal approval rather than genuinely addressing financial inclusion and social responsibility.



# Proposed Stakeholder-Legitimacy Framework

In the early 2000s, @HOOGHIEMSTRA2000 argues that sustainability reporting research exhibits diverse and inconsistent findings, primarily due to the absence of a comprehensive theoretical framework. The research asserts that legitimacy was the dominant perspective. @SPENCE2010 identified stakeholder theory as the predominant and most effective framework for explaining sustainability reporting practices. They also point out that while many studies mention stakeholders in general, they do not explicitly reference stakeholder theory or other theoretical frameworks. Our review confirms their observations in the literature of banks' ESG disclosure. The majority of the studies show a preoccupation with stakeholder theory [@GALANT2017; @SHEN2016; @BUALLAY2021], legitimacy theory [e.g. @CARNEVALE2014], and to some extent also institutional theory [@HIGGINS2014; @BEBBINGTON2018; @CHRISTENSEN2021]. Moreover, these studies primarily rely on isolated theoretical frameworks rather than adopting a more holistic approach that integrates multiple theoretical perspectives on ESG disclosure. In this paper we propose an integrated Stakeholder-Legitimacy Framework of banks' ESG disclosure, which synthesises and extends insights from @CAMPBELL2007 institutional-economic model and @AGUINIS2012 multilevel CSR analysis. At its core, the framework positions ESG disclosures as outcomes of dynamic negotiations between stakeholder pressures and legitimacy-seeking behaviours, moderated by banks' economic and institutional contexts. 


```{r stakeholder-legitimacy framework}
#| label: fig-flowchart
#| fig-cap: "The Stakeholder-Legitimacy Framework for ESG Disclosure in Banking"

library(DiagrammeR)
library(DiagrammeRsvg)
library(rsvg)

grViz("
digraph {
  graph [
    rankdir = TB,
    size = '8,10!',        // Constrain graph dimensions
    ratio = 'compress',    // Compress to fit dimensions
    nodesep = 0.5,         // Reduce spacing between nodes
    ranksep = 0.4,         // Reduce vertical spacing
    fontsize = 8          // Reduce font size
  ]
  
  node [
    shape = oval,
    width = 1.5,          // Reduce node width
    height = 0.8,         // Reduce node height
    fontname = 'Helvetica'
  ]
  
  // Main vertical flow
  1 [label = 'Stakeholder Pressure\n(Whom banks prioritise)']
  2 [label = 'Legitimacy-seeking Behaviour\n(Why banks disclose)']
  3 [label = 'ESG Disclosure']
  
  // Side nodes
  4 [label = 'Bank Economic Context\n(size, performance)']
  5 [label = 'Bank Institutional Context\n(ownership, regulation)']
  
  // Main vertical connections
  1 -> 2 -> 3 [minlen = 1]
  
  // Side connections
  4 -> 3 [constraint = false]
  5 -> 3 [constraint = false]
  
  // Force side nodes to same rank
  {rank = same; 4; 5}
  
  // Position side nodes using invisible edges
  invisible_node [shape = point, width = 0, height = 0]
  2 -> invisible_node [style = invis]
  invisible_node -> 4 [style = invis]
  invisible_node -> 5 [style = invis]
}
") %>% 
  export_svg() %>% 
  charToRaw() %>% 
  rsvg_pdf("flowchart.pdf")

knitr::include_graphics("flowchart.pdf")

```


In the integrated Stakeholder-Legitimacy framework (@fig-flowchart), stakeholder pressures act as catalysts for banks' ESG disclosures. According to stakeholder theory, companies should consider and balance the diverse viewpoints and expectations of all groups affected by or invested in their operations, rather than focusing solely on shareholders [@BUCHHOLZ2005; @LAPLUME2008]. @FREEMAN1984 suggests that company management should stay attuned to changing dynamics and trends affecting both their organisation's internal constituents and outside parties. Stakeholder pressures arise from key external actors, including regulators, investors, and customers. These actors demand greater transparency and accountability, motivating banks to provide valuable explanations of how they answer to the societal call for sustainable business conduct. Banks respond to these pressures through strategic legitimacy-seeking behaviours. Banks' ESG disclosures range from symbolic gestures (e.g. adopting TCFD guidelines without operational changes) to substantive actions such as phasing out coal financing. Crucially, legitimacy is not a static achievement but an ongoing process; banks must continuously adapt to shifting norms, such as the rise of the TNFD as a biodiversity benchmark. In the stakeholder-legitimacy context, influential stakeholders execute greater pressure on companies to explain and justify their business conduct. Therefore, sustainable reporting and disclosed ESG information serve as a way for companies to establish and maintain their legitimacy in the eyes of these stakeholders [@CAMPBELL_D2003]. 

The integrated framework further incorporates moderating contexts that shape ESG disclosures produced by the translation of stakeholder pressures into banks' legitimacy-seeking behaviours, including economic conditions such as size and profitability, and institutional context such as ownership structure and regulatory regimes. These contexts fundamentally determine banks' capacity and incentives to pursue substantive versus symbolic legitimacy responses​ [@CAMPBELL2007; @AGUINIS2012]. 

The framework incorporates two primary dimensions of moderating context.​​ The ​economic context, primarily manifested in bank size and financial performance, influences the resources available and the cost-benefit calculus of ESG engagement. For example, Larger size typically affords ​greater visibility​ [@MEZNAR1995] ​and resource advantages​ [@YOUN2015; @FAVINO2019], while stronger financial performance provides the ​necessary resource slack​ [@HADDOCK2005; @LIU2009; @AGUINIS2012] to absorb ESG costs. Concurrently, the ​institutional context, encompassing factors such as ownership structure (including listing status, institutional shareholding, state ownership, and ownership concentration) and the stringency of the regulatory environment, critically alters ​governance priorities, stakeholder salience​ [@MITCHELL1997], and the ​legitimacy thresholds​ banks must cross to justify their ESG actions. In particular, the regulatory environment ​codifies stakeholder expectations into enforceable norms​ [@HESS2007; @DESAI2024], defining the boundaries between symbolic compliance and substantive accountability. Consequently, these moderating factors do not merely correlate with disclosure levels; they ​actively shape the strategic choices​ banks make in translating stakeholder pressures into legitimacy-seeking ESG disclosure behaviours within specific economic and institutional constraints [@CAMPBELL2007; @AGUINIS2012].


# Conclusion

The transition from voluntary ESG initiatives to mandatory disclosure frameworks marks a paradigm shift in how financial institutions conceptualize and manage nature-related risks. Global frameworks like the TNFD and ISSB, alongside regional regulations such as the EU Taxonomy and CSRD, have established rigorous reporting standards that integrate environmental impacts into financial decision-making. For banks, these mandates necessitate not only transparency about their own operations but also heightened scrutiny of borrowers’ sustainability practices [see @WANG2023], creating cascading effects across economies. Empirical evidence suggests that ESG disclosures can mitigate bank-specific risks—such as insolvency and liquidity challenges—while improving market efficiency through reduced information asymmetry [@GANGI2019; @SCHOLTENS2019; @GANGWANI2024]. However, conflicting findings on the financial performance-ESG relationship and the role of ownership structures underscore the need for contextualized analyses.

To address these complexities, this study advances the Stakeholder-Legitimacy Framework and tailor it to the banking sector. This framework synthesises stakeholder theory, legitimacy theory, and institutional isomorphism to explain how banks’ ESG disclosures emerge from dynamic negotiations between external pressures (e.g., regulators, investors) and legitimacy-seeking behaviours. It posits that banks balance competing stakeholder demands—such as investor calls for transparency and regulatory mandates for biodiversity due diligence—through strategic disclosures that range from symbolic compliance (e.g., superficial TNFD adoption) to substantive action (e.g., divesting from deforestation-linked assets). Crucially, the framework incorporates moderating factors such as ownership structures (e.g., institutional ownership, state ownership) and regulatory regimes (e.g., mandatory CSRD, voluntary IFRS S1 and S2 adoption), offering a nuanced lens to analyze how economic and institutional contexts shape disclosure outcomes [@CAMPBELL2007; @AGUINIS2012]. 

While this framework provides a robust analytical tool, important research gaps warrant attention. Key areas include inconsistent findings regarding ownership structures' influence on ESG outcomes [see @DYCK2019; @ALUCHNA2022], insufficient exploration of how mandatory ESG regulations shape bank risk profiles in practice [see @WANG2023], and banks' dual role as both reporting entities and sustainability gatekeepers [see @SMITH2022]. Additionally, mixed evidence on the ESG-financial performance nexus highlights the need for context-specific analyses [see @BORGES2024].

Future research may prioritise three avenues: (1) disentangling the economic and institutional factors that drive ESG disclosure quality in banks; (2) evaluating the systemic risk implications of nature-related regulatory shocks; for instance, assess how nature-related disclosures influence systemic risk within the banking sector and the wider economy; and (3) exploring the interaction between stakeholder-legitimacy framework and institutional isomorphism framework, to assess how coercive, mimetic, and normative pressures shape banks' ESG disclosure practices. 

The proliferation of nature-related disclosure frameworks in recent years reveals the shift from voluntary toward mandatory disclosure: from TNFD's market-driven recommendations to IFRS S1 and S2 disclosure framework, and mandatory EU reporting standards CSRD. This trend is driven by several key developments including recognition of material nature-related risks, global reporting alignment, and integration with financial reporting. However, accelerating fragmentation creates compliance complexity, while implementation gaps persist in emerging economies facing data scarcity and capacity constraints. The fourth research direction may therefore focus on: harmonization pathways across frameworks, effectiveness of disclosure mandates in redirecting conservation financing (e.g. how metric inconsistencies create greenwashing vulnerabilities), and tangible impacts on biodiversity loss reduction—particularly in high-risk lending portfolios.​

By focusing on these research avenues through the lens of the Stakeholder-Legitimacy Framework, scholars and policymakers can enhance our understanding of ESG disclosure dynamics in the banking sector. This will not only refine regulatory architectures but also ensure that ESG disclosures go beyond compliance checklists to drive meaningful progress. 



# Appendix: Regional ESG Reporting Framework Table

\newpage

```{r KableExtra-Regional Table}
#| label: tbl-appendix
#| tbl-cap: "Appendix: Regional ESG Disclosure Regulations"
#| echo: false

library(tidyverse)
library(kableExtra)

## Create variables
Region <- c("EU",
            "EU", 
            "EU", 
            "UK", 
            "US", 
            "US")

Regulation <- c("EU Taxonomy Regulation", 
                "Corporate Sustainability Reporting Directive (CSRD)", 
                "Sustainable Finance Disclosure Regulation (SFDR)", 
                "Sustainability Disclosure Requirements (SDR)", 
                "SEC Climate Disclosure Rule", 
                "Climate Corporate Data Accountability Act (SB 253)")

Year <- c(2020, 
          2023, 
          2021, 
          2023, 
          2024, 
          2026)

Key_features <- c("a comprehensiver classification system designed to identify environmentally sustainable activities", 
                  "'double materiality' approach; ESRS", 
                  "financial institutions classify investment products based on their sustainability objectives", 
                  "ISSB-aligned reporting; Transition plans for Paris and GBF; Financed emissions disclosure", 
                  "Material climate risk disclosure; Scope 1-3 emissions reporting; Extreme weather cost reporting",
                  "Scope 1-3 emissions reporting for $1B+ revenue firms; Biodiversity risk assessment")
  
Scope <- c("Banks and in-scope entities", 
           "50,000+ EU/non-EU companies by 2027", 
           "Financial institutions", 
           "Listed companies (from 2026); High-impact sectors (agriculture, mining)", 
           "Public companies", 
           "Large corporations operating in California")

Enforcement <- c("Failure to comply can result in reputational damage and exclusion from sustainable finance markets", 
                 "Mandatory for EU/non-EU companies", 
                 "Market exclusion for non-compliance", 
                 "FCA enforcement starting 2026", 
                 "Delayed due to legal challenges", 
                 "California state law enforcement")

Alignment <- c("EU Taxonomy classification system", 
               "Aligns with IFRS S1 and S2 standards; TNFD metrics", 
               "Links to EU Taxonomy", 
               "Aligns with IFRS S1 and S2; References TNFD recommendations", 
               "N/A", 
               "Addresses supply chain biodiversity risks")
  
## Set up the tibble
df_region <- tibble(Region, Regulation, Year, Key_features, Scope, Enforcement, Alignment)

kableExtra::kbl(df_region, 
                booktabs = TRUE) |> 
  kable_styling(latex_options = c("striped", "scale_down"), 
                font_size = 11) |> 
  row_spec(row = 0, bold = TRUE, background = "lightgray", align = "c") |> 
  column_spec(1, width = "5em") |>     # Region
  column_spec(2, width = "8em") |>   # Regulation
  column_spec(3, width = "5em") |>   # Year
  column_spec(4, width = "12em") |>     # Key_Requirements
  column_spec(5, width = "8em") |>   # Scope
  column_spec(6, width = "10em") |>   # Enforcement
  column_spec(7, width = "10em")  
 
```


\newpage




















