---
title: "A Stakeholder-Legitimacy Framework for ESG Disclosure in Banking Under Evolving Regulatory Mandates"
date: today
dateformat: long
author: 
  - name: Ying (Veronica) Zhang
    email: veronica.zhang@qub.ac.uk
    affiliations: 
      - id: sbs
        name: Queen's University Belfast
        department: Institute for Global Food Security (IGFS) School of Biological Sciences
        address: Belfast, UK
    attributes:
      corresponding: true
    note: |
      This work was supported by the BIOFIN-EU Project (Grant agreement 101135476, UKRI through the UK Horizon Guarantee Fund (no.10092956) and through the National Environment Research Council (Grant no. Z503368)  
  - name: Lisa Sheenan
    affiliations: 
      - id: ucd
        name: University College Dublin
        department: UCD Michael Smurfit Graduate Business School
        address: Dublin, Ireland
      - ref: sbs
  - name: Fearghal Kearney
    affiliations: 
      - id: qbs
        name: Queen's University Belfast
        department: Queen's Business School
        address: Belfast, UK
  - name: Junyu Zhang
    affiliations: 
      - ref: sbs  
  - name: George Hutchinson
    affiliations: 
      - ref: sbs
  - name: Neil Reid
    affiliations: 
      - ref: sbs
      - id: cbw
        name: Queen's University Belfast
        department: Co-Centre for Climate + Biodiversity + Water, School of Biological Sciences
        address: Belfast, UK
  - name: Paul Brereton
    affiliations: 
      - ref: sbs          

abstract: |
  The global financial sector is undergoing a transformative shift as ​material risks stemming from environmental and social challenges, such as​ biodiversity loss, ecosystem degradation, ​and climate change,​​ underscore the growing importance of Environmental, Social, and Governance (ESG) factors for economic stability. By reviewing the literature, this study advances a novel Stakeholder-Legitimacy Framework, integrating stakeholder theory, legitimacy theory, and institutional isomorphism to explain how banks’ ESG disclosures emerge from dynamic negotiations between external pressures (e.g., regulators, investors) and strategic legitimacy-seeking behaviors. The study reveals that while ESG disclosures mitigate bank-specific risks, conflicting evidence persists on their financial performance impacts. Critical gaps include underexplored transmission effects of regulations through lending networks, inconsistent roles of institutional and state ownership, and tensions between symbolic compliance and substantive action.
keywords: 
  - ESG disclosure
  - IFRS S1 and S2
  - Stakeholder-Legitimacy framework
  - bank risk
  - ownership
  - climate change
  - JEL code M14, M41, G28
bibliography: reviewurl.bib
format: 
  elsevier-pdf: 
    keep-tex: true
    header-includes: |
      \usepackage{graphicx}
      \usepackage{adjustbox}
    journal: 
      name: "The British Accounting Review"
      cite-style: authoryear
    geometry: [left=3.5cm, right=3.5cm, top=4.5cm, bottom=4.5cm]
execute: 
  eval: true
  echo: false
  output: true
  warning: false
  error: false
  include: true
---

```{r set up}
#| label: libraries
#| output: false
#| include: false

library(tidyverse)
library(kableExtra)
library(DiagrammeR)
library(DiagrammeRsvg)
library(rsvg)
library(knitr)
library(tinytex)




```

\newpage

# Introduction

The global financial system is undergoing a seismic shift as nature-related risks—ranging from biodiversity loss to ecosystem degradation—emerge as material threats to economic stability [@UNEP2023]. Over the past decade, voluntary sustainability frameworks like the Task Force on Climate-related Financial Disclosures (TCFD) have evolved into robust, jurisdictionally mandated systems, driven by regulatory imperatives to combat greenwashing, harmonise reporting standards, and align financial flows with planetary boundaries. Central to this transformation are frameworks such as the Task Force on Nature-Related Financial Disclosures (TNFD), the International Sustainability Standards Board (ISSB), and region-specific regulations like the EU’s Corporate Sustainability Reporting Directive (CSRD) and the UK’s Sustainability Disclosure Requirements (SDR). These frameworks reflect a growing consensus that environmental risks are inextricably linked to financial performance, necessitating disclosures that integrate double materiality—assessing both financial risks to firms and firms’ impacts on ecosystems.

Banks are uniquely exposed to systemic ESG risks due to their role as financial intermediaries, warranting specialized regulatory analysis. Nevertheless, empirical research on how ESG rules influence their risk exposure, governance, and performance remains fragmented. Existing studies disproportionately focus on non-financial firms [e.g. @YOUN2015; @FAVINO2019; @BUALLAY2019; @DYCK2019], leaving critical gaps in understanding banks’ unique intermediation role. For instance, while the TNFD’s LEAP methodology and the ISSB’s IFRS S1 and S2 standards are gaining traction, their sector-specific implications for banking—such as transmission effects through lending networks or the tension between symbolic compliance and substantive action—are underexplored. Theoretical frameworks also remain fragmented, with stakeholder theory, legitimacy theory, and institutional isomorphism often treated in isolation rather than as interconnected drivers of ESG disclosure.[^01] This paper addresses these gaps by synthesising empirical evidence, proposing an integrated theoretical model, and identifying urgent research priorities at the nexus of finance, regulation, and ecological stewardship.

[^01]: Stakeholder theory [@FREEMAN1984; @MITCHELL1997], legitimacy theory [@SUCHMAN1995], and institutional isomorphism [@DIMAGGIO1983] have been applied individually to explain ESG disclosure motivations. Yet, to our knowledge, ​no major study​ has synthesises these three frameowrks to examine their synergetic effects, creating a theoretical silo problem. This fragmentation mirrors broader critiques of theoretical dis-integration in ESG disclosure [@ECCLES2012].

Current empirical studies show how banks use ESG reporting to navigate real-world constraints. @THOMPSON2004 examines how banks respond to environmental stakeholders through reporting practices. Their findings reveal that bankers value annual reports for assessing corporate environmental impact, despite their limitations, and express some interest in expanded environmental disclosures. However, their focus primarily aims at safeguarding loans. Their study reflects tensions between stakeholder demands and bank risk management. Similarly, @SCHOLTENS2009 analyze how international banks balance different stakeholder interests through Corporate Social Responsibility (CSR) reporting, while @CARNEVALE2014 test the direct effects of the sustainability report on stock price in European markets, demonstrating that sustainability reports positively influence stock prices by signaling organizational compliance with evolving norms. @MURE2021 provide empirical evidence supporting stakeholder theory and legitimacy theory that ESG disclosure and performance can be used by banks as a tool to manage their reputation and maintain stakeholder trust, especially in the face of adverse events. These findings underscore a critical gap: Existing theories fail to fully explain how banks reconcile competing stakeholder, legitimacy, and economic imperatives in ESG implementation. This necessitates the integrated Stakeholder-Legitimacy Framework proposed in Section IV.  

In this study, we advance a novel Stakeholder-Legitimacy Framework [@CAMPBELL2007; @AGUINIS2012] tailored to banks, synthesising stakeholder theory, legitimacy theory, and institutional isomorphism. This model positions ESG disclosures as dynamic outcomes of external pressures (e.g., regulators, shareholders) and banks’ legitimacy-seeking behaviors, moderated by economic factors (e.g., bank size) and institutional contexts (e.g., ownership structures). Our synthesis of extant literature highlights three critical research gaps:

- Ownership Dynamics: The role of institutional and state ownership in shaping ESG outcomes remains inconsistent across studies [@DYCK2019; @RAIMO2020], with concentrated ownership marginalising minority shareholder influence on CSR practices [@SMITH2022].

- Regulatory Efficacy: Sector-specific analyses of mandatory frameworks (e.g., CSRD, EU Taxonomy) are scarce, particularly their impact on bank risk profiles and governance mechanisms.

- Systemic Risk Implications: The interplay between ESG disclosures and systemic financial stability, especially during regulatory shocks (e.g., rapid green  and climate transitions), lacks empirical rigor [@ESRB2016].

Our study contributes to sustainable finance research through three key areas:

1. Theoretical Innovation: The proposed Stakeholder-Legitimacy Framework bridges siloed theories, offering a holistic lens to analyze banks’ ESG disclosures as context-dependent processes shaped by stakeholder pressures and legitimacy imperatives. Addressing a limitation of theory isolation [@ECCLES2012], our framework conceptualises banks’ ESG disclosures as dynamic outcomes of external stakeholder pressures (e.g., regulators, activist investors) and internal legitimacy-seeking behaviours, moderated by institutional contexts such as ownership structures. 

2. Gap-Driven Research Agenda: we identify and critically assess two underexplored dimensions of banking sector ESG research. (1) Regarding ownership dynamics, empirical inconsistencies and state ownership meditate CSR outcomes [@DYCK2019; @RAIMO2020]. (2) Regarding regulatory efficacy, sector-specific analyses of mandatory frameworks (e.g., CSRD, EU Taxonomy) are scarce, particularly their capacity to influence bank risk profiles and governance mechanisms [@ESRB2016]. By mapping these gaps, this study establishes a targeted agenda to address banking-specific complexities.

3. Policy Relevance: Our analysis yields actionable recommendations for regulators and polocymakers, including integrating natural capital valuation into stress testing and harmonizing ISSB-TNFD reporting interoperability to reduce compliance arbitrage opportunities. These proposals respond to mounting concerns about symbolic compliance in sustainability reporting [@SMITH2022], offering concrete steps to align regulatory frameworks with planetary boundary imperatives.

The remainder of the literature review is organized as follows: In Section II and III, we review the theories underpinning, and the empirical studies on the relationship between ESG disclosure and the banking sector, respectively. In Section IV, we propose the integrated Stakeholder-Legitimacy framework applied to the banking industry and review the research gaps. Section V is the conclusion section. An overview of the nature-related financial disclosure frameworks from different dimensions, as background information of this review, is provided in the appendix. 


# Theoretical Foundation

The relationship between environmental, social, and governance (ESG) reporting and financial institutions has garnered increasing scholarly attention, driven by global sustainability challenges. This section first establishes the conceptual landscape of contemporary ESG disclosure before synthesizes theoretical frameworks that explain how and why ESG reporting influences financial institutions’ operations, decision-making, and market outcomes. 


## Conceptual Advances in ESG Disclosure

Contemporary theoretical concepts and analytical lenses around natural capital and double materiality offer novel perspectives on integrating environmental considerations into financial decision-making. ESG reporting is not merely a compliance exercise but a strategic imperative for financial institutions. Natural capital concept, as articulated by @DASGUPTA2021 and @DAILY2009, emphasizes the valuation of biodiversity and natural capital, with implications for financial modeling [@ATKINSON2014]. @ECCLES2014 provide a theoretical foundation for integrated reporting,  which aligns closely with double materiality. The concept of "double-materiality" was first formally proposed by the European Commission [@EC2019] in Guidelines on Non-financial Reporting: Supplement on Reporting Climate-related Information published in June 2019; then was formalized by the regulation of the Corporate Sustainability Reporting Directive (CSRD)[see @EU_CSRD_2022] as a reporting requirement. It mandates a company to judge materiality from two perspectives 1) “the extent necessary for an understanding of the company’s development, performance and position” and “in the broad sense of affecting the value of the company”; 2) environmental and social impact of the company’s activities on a broad range of stakeholders. 


## Organizational Theories

Organizational theories, particularly stakeholder theory and legitimacy theory, highlight the institutional motivations and external pressures driving ESG reporting practices.  Stakeholder theory, developed by @FREEMAN1984, posits that organizations are accountable to a broad range of stakeholders beyond shareholders, including regulators, customers (investors), employees, and civil society. Building on this, @MITCHELL1997 proposed the theory of stakeholder salience emphasizing understanding why and how managers prioritize certain stakeholders, rather than prescribing whom they should prioritize. Their theory provides tools to identify stakeholder types via a typology and explain managerial responses through the concept of salience, which assesses stakeholders' legitimacy, power, and urgency. This typology offers a framework for financial institutions to navigate competing stakeholder interests effectively. @DONALDSON1995 provided theoretical justification for stakeholder theory, emphasizing its descriptive, instrumental, and normative validity in explaining corporate behavior. @ROBERTS1992 conducts an application of stakeholder theory to explain corporate social responsibility disclosure. The study finds that measures of stakeholder power, strategic posture, and economic performance are significantly associated with levels of corporate social disclosure. Banks' ESG disclosure, for instance, on climate risks or green lending portfolios are often framed as responses to investor demands for transparency or regulatory mandates like the EU Taxonomy and the EU’s Corporate Sustainability Reporting Directive (CSRD). 

Legitimacy theory, foundationally developed by @SUCHMAN1995, identifies three primary forms of legitimacy[^03]—pragmatic, moral, and cognitive—that organizations seek to maintain. @DEEGAN2002 applied this framework to environmental and social disclosures, demonstrating how these practices reinforce an organization's legitimacy. Firms' legitimacy-seeking behaviors influence the depth and scope of disclosures, with firms in environmentally sensitive industries disclosing more to manage public perception. In banking contexts, ESG disclosures function as legitimacy-management tools to address stakeholder pressures, as evidenced by sector-specific studies [see @THOMPSON2004; @MURE2021; @CARNEVALE2014].  

Building on stakeholder theory and legitimacy theory, institutional isomorphism elucidates the structured mechanisms that drive organizational conformity. @DIMAGGIO1983 identify three isomorphic drivers that create organizational homogeneity:

1. **Coercive isomorphism**: explains how regulatory requirements and pressure from stakeholders drive adoption of standardized reporting formats. 

2. **Mimetic isomorphism**: illuminates why organizations copy "best practice" reporting methods from industry leaders during periods of uncertainty. 

3. **Normative isomorphism**: shows how professional networks and training lead to similar reporting approaches across organizations.

Their foundation study develops a theoretical framework explaining how these mechanisms operate within organizational fields. This framework is valuable in understanding the standardization and diffusion of reporting practices across organizations. Empirical research highlights the importance of institutional pressures, including coercive, normative, and mimetic forces, in driving firms to conform to ESG expectations. @DELMAS2004 demonstrate firms' heterogeneous interpretations of isomorphic forces determine substantive versus symbolic ESG integration. This divergence mirrors banking practices where identical EU CSRD mandates yield divergent TNFD implementation depth. @BEBBINGTON2018 position normative isomorphism as instrumental in bridging accounting praxis with planetary boundaries, arguing that professional networks institutionalize "double materiality" reporting, which notes the field's underdevelopment in translating SDGs into bank capital allocation. @CHRISTENSEN2021 reveal mimetic isomorphism's unintended consequences: firms under US/EU disclosure mandates increasingly adopt boilerplate ESG language to signal conformity while obscuring institution-specific ecological impacts.  


## Economic Theories

Economic theories, particularly information asymmetry and agency theory, explain how ESG reporting enhances market efficiency and reduces capital costs. @JENSEN1976 established the foundational agency theory framework, explaining the conflicts of interest between principals (shareholders) and agents (managers). This study lays the groundwork for understanding how ESG reporting can mitigate agency costs by aligning stakeholder interests. @AKERLOF1970 work on Information Asymmetry further explains how unequal information among market participants creates inefficiencies. @HART1987 provide a foundation for modern contract theory and influence the theory of the firm and organizational economics. This paper develops the principal-agent model, and demonstrate how disclosure requirements can act as screening devices. @GRAY1996 provide a theoretical framework for understanding the role of corporate social and environmental reporting in addressing information asymmetry and enhancing accountability. Applying these theories, studies have demonstrated that ESG disclosure reduces information gaps and lowers agency costs, thereby enhancing market efficiency. For instance, @DHALIWAL2011 and @ELGHOUL2011 find firms that initiate CSR disclosure and have superior CSR performance (relative to their industry peers) experience a subsequent reduction in their cost of equity capital, and attract more dedicated institutional investors and analyst coverage. Moreover, @CUI2018 observed that CSR engagement is inversely associated with reputation risk, largely due to reduced information asymmetry. 

[^03]: Legitimacy is a generalized perception or assumption that the actions of an entity are desirable, proper, or appropriate within some socially constructed system of norms, values, beliefs, and definitions [@GINZEL2004; @NEILSEN1987; @PERROW1970].


# Extant Empirical Evidence

## Firm Value and Performance

Stakeholder theories [see @FREEMAN1984; @DEEGAN2002] suggest that when organizations incorporate environmental, social, and governance (ESG) practices into their long-term strategic planning, they gain competitive advantages by addressing the needs of various groups - from employees and customers to government bodies and community members. This approach recognizes that business success is linked to creating value for all stakeholders, not just shareholders. The stakeholders reward companies with good ESG practices through investment, consumption and higher productivity [@LI2018]In contrast, the trade-off hypothesis or traditionalist view [see @FRIEDMAN2007] suggests ESG practices negatively impact financial performance by increasing costs, harming corporate performance, and reducing competitive advantage. Scholars argue that focusing on social and environmental goals diverts managers from maximizing shareholder value. Similarly, satisfying non-shareholder stakeholders may hinder profit maximization and value creation for owners and managers[@GALANT2017].

Empirical studies on the relationship between ESG engagement and financial performance have provided conflicting findings. Driven by stakeholder theory, @WU2013 find that ESG disclosures improve bank profitability by attracting ESG-conscious investors. Their findings attribute to stakeholder pressure for transparency. Drawing from empirical analysis using MSCI ESG STATS database ratings, @CORNETT2016 examine the relationship between the US commercial banks' corporate social responsibility (CSR) practices and financial performance. The study reveals that larger banks, despite facing criticism for profit-focused practices leading to the crisis, demonstrate consistently higher CSR strengths and concerns. Post-2009, these institutions showed marked improvement with increased CSR strengths and decreased concerns. Similarly, studies such as @CARNEVALE2014, @SHEN2016 and @BUALLAY2021 demonstrate banks' engagement in CSR activities can improve the financial performance. @TIAN2023 report that banks with green credit announcement show better market value and long-term performance. While @FERRERO2016 find a nonlinear relationship between ESG and financial performance employing listed companies in Europe. Moreover, @BUALLAY2023 investigate the effect of sustainability reporting on bank performance (operational, financial and market) in 7 regions that include 60 countries over the period 2008-2017. They find the negative relationship between ESG disclosures and banks' operational performance (ROA, ROE), and market performance (Tobin's Q). They argue that sustainability reporting may have negative impact on banks' asset utilization, in line with the trad-off theory [see @LEE2009]. @ARAS2024 investigates the SDGs with ESG indicators through a double materiality perspective for 1888 companies from the OECD financial institutions. The study shows how commercial banks can identify and prioritize the SDGs and targets and how sustainable practices at the corporate level can contribute to achieving these global goals by adopting a sound approach.


## Bank Risk 

Despite growing scholarly attention to the link between CSR and bank performance, empirical research examining its influence on risk exposure within financial institutions remains limited. The existing literature primarily focuses on individual bank risk metrics. A study by @GANGI2019 examine 142 banks across 35 countries and find that stronger environmental engagement correlates with lower risk profiles as measured by Z-scores [see @LAEVEN2009]. Similarly, @SCHOLTENS2019 report a modest negative relationship between sustainability practices and standalone risk measures among European banks. Aligning with stakeholder theory, @DI_TOMMASO2020 demonstrate that higher ESG scores slightly curbed risk-taking among European banks (2007–2018), though this relationship was mediated by board composition factors including size, independent director presence, and gender diversity. @GANGWANI2024 report that higher ESG disclosure scores correlate with reduced commercial banking risks, specifically mitigating insolvency, leverage, and liquidity challenges. Further evidence from @CHIARAMONTE2022 demonstrate that both aggregate ESG scores and individual pillars enhanced bank stability during financial stress periods, with longer ESG disclosure histories amplifying these stabilizing effects. Expanding geographically, @NEITZERT2022 examine 582 international banks (2002–2018), corroborating the risk-mitigating effects of CSR activities on default and portfolio risks, consistent with risk management frameworks. However, their findings revealed heterogeneity across ESG pillars: environmental (E) initiatives exhibited the strongest and most consistent risk-reduction outcomes, whereas social (S) and governance (G) dimensions yielded less definitive results.

Academic research on the relationship between systemic financial risk and ESG factors remains limited, with existing studies examining each of the three E, S and G pillars respectively. Regarding environmental (E) factors, @ESRB2016 points that an unanticipated rapid green transition could significantly impact asset values, with carbon-intensive investments facing declining profitability while low-carbon assets appreciate. In terms of governance (G) factors, @ANGINER2018 present compelling evidence that shareholder-friendly corporate governance structures correlate with increased bank systemic risk, as measured by Marginal Expected Shortfall (MES)[@ACHARYA2017] and SRISK metrics [@BROWNLEES2017].This relationship appears particularly pronounced for larger banks, potentially due to too-big-to-fail considerations, and is amplified in jurisdictions with more generous financial safety nets. Their findings derive from analyses of both U.S. banks (1990-2014) and international banks (2004-2008), suggesting the robustness of this relationship across different contexts and time periods.  

Using qualitative research method, @KUHN2022 investigates sustainable finance disclosures and initiatives in Germany. They find that ESG disclosures enable financial institutions to identify exposure to sectors with high dependency on vulnerable ecosystems. Disclosing these dependencies improves transparency and helps institutions avoid investments in activities contributing to ecological degradation such as deforestation-linked loans. Their study proposes integrating natural capital valuation into credit risk assessments to account for "hidden" risks. Case studies demonstrate that institutions adopting such frameworks reduce exposure to stranded assets and regulatory penalties.

The literature on the relationship between ESG factors and bank risk is summarised in @tbl-literature.

```{r literature table}
#| label: tbl-literature
#| tbl-cap: "Summary of Literature on ESG and Financial Institutions"

df <- tibble(
  "Risk Type" = c("Individual Bank Risk", 
                  "Individual Bank Risk", 
                  "Individual Bank Risk", 
                  "Systemic Financial Risk", 
                  "Systemic Financial Risk", 
                  "Qualitative Risk Insights"
                  ), 
  "Findings" = c("Higher ESG scores linked to lower insolvency, leverage, and liquidity risk (e.g Gangi et al., 2019)",
                  "Longer ESG reporting history enhances risk mitigation (Chiaramonte et al., 2022)",
                  "Environmental (E) pillar shows strongest risk-reduction effect (Neizert and Petras, 2022)", 
                  "Rapid green transition can destabilise assets (ESRB, 2016)", 
                  "Shareholder-friendly governance may increase systemic risk (Anginer et al., 2018)", 
                  "Natural capital integration may reduce exposure to ecological risks (Kuhn, 2022)"
                  ), 
  "Implication" = c("Supports ESG as a risk management tool", 
                    "Suggests consistency and time horizon of disclosure matters", 
                    "Emphasis importance of environmental reporting", 
                    "Calls for climate transition scenario analysis", 
                    "Raises concerns for governance-driven risk in large banks", 
                    "Points to improved credit risk assessment via ESG reporting"
                  ) 
)


df_kbl <- kbl(df, 
              booktabs = TRUE, 
              format = "latex") |>
          kable_styling(latex_options = c("scale_down"), 
                        font_size = 12) |>
          row_spec(0, bold = TRUE) 

df_kbl

```



## Bank Lending, ESG Risk and Strategic Alignment

Banks have revised their approach to corporate social responsibility (CSR), placing greater emphasis on managing both direct and indirect risks associated with lending to firms facing environmental and social challenges [@CARNEVALE2012]. Drawn on the stakeholder theory, @GOSS2011 examine the link between CSR and the cost of bank loans. They find that firms with poor CSR performance face higher loan costs due to the creditor risk perceptions. 

@DEMETRIADES2025 investigate lending patterns between major banks and fossil fuel companies from 2001-2021 using global syndicated loan data. They find a complex dynamic where banks recognize and price in climate risks through higher interest rates and shorter loan terms, yet simultaneously increase loan volumes to brown firms. The findings suggest that regulatory pressure, particularly in Europe and the US, has led to more stringent lending policies, though not necessarily reduced lending volumes.

@BASU2022 report that high-ESG banks complement, rather than substitute, mortgage lending with community development investments in poor areas. However, banks are more likely to reject mortgage applications from these communities, suggesting social washing—using pro-social rhetoric while limiting actual support. Their findings align with the legitimacy theory, as banks appear to engage in CSR initiatives to maintain societal approval rather than genuinely addressing financial inclusion and social responsibility.



# Proposed Stakeholder-Legitimacy Framework

## The Theory  

In the early 2000s, @HOOGHIEMSTRA2000 argues that sustainability reporting research exhibits diverse and inconsistent findings, primarily due to the absence of a comprehenCHsive theoretical framework. The research asserts that legitimacy was the dominant perspective. @SPENCE2010 identified stakeholder theory as the predominant and most effective framework for explaining sustainability reporting practices. They also point out that while many studies mention stakeholders in general, they do not explicitly reference stakeholder theory or other theoretical frameworks. Our review confirms their observations in the literature of banks' ESG disclosure. The majority of the studies show a preoccupation with stakeholder theory [@GALANT2017; @SHEN2016; @BUALLAY2021], legitimacy theory [e.g. @CARNEVALE2014], and to some extent also institutional theory [@HIGGINS2014; @BEBBINGTON2018; @CHRISTENSEN2021]. Moreover, these studies primarily rely on isolated theoretical frameworks rather than adopting a more holistic approach that integrates multiple theoretical perspectives on ESG disclosure. In this paper we propose an integrated Stakeholder-Legitimacy Framework of banks' ESG disclosure, which synthesizes and extends insights from @CAMPBELL2007 institutional-economic model and @AGUINIS2012 multilevel CSR analysis. At its core, the framework positions ESG disclosures as outcomes of dynamic negotiations between stakeholder pressures and legitimacy-seeking behaviors, moderated by banks' economic and institutional contexts. 


```{r stakeholder-legitimacy framework}
#| label: fig-flowchart
#| fig-cap: "ESG Stakeholder-Legitimacy Disclosure Framework"

library(DiagrammeR)
library(DiagrammeRsvg)
library(rsvg)

grViz("
digraph {
  graph [
    rankdir = TB,
    size = '8,10!',        // Constrain graph dimensions
    ratio = 'compress',    // Compress to fit dimensions
    nodesep = 0.5,         // Reduce spacing between nodes
    ranksep = 0.4,         // Reduce vertical spacing
    fontsize = 8          // Reduce font size
  ]
  
  node [
    shape = oval,
    width = 1.5,          // Reduce node width
    height = 0.8,         // Reduce node height
    fontname = 'Helvetica'
  ]
  
  // Main vertical flow
  1 [label = 'Stakeholder Pressure\n(Whom banks prioritise)']
  2 [label = 'Legitimacy-seeking Behaviour\n(Why banks disclose)']
  3 [label = 'ESG Disclosure']
  
  // Side nodes
  4 [label = 'Bank Economic Context\n(size, performance)']
  5 [label = 'Bank Institutional Context\n(ownership, regulation)']
  
  // Main vertical connections
  1 -> 2 -> 3 [minlen = 1]
  
  // Side connections
  4 -> 3 [constraint = false]
  5 -> 3 [constraint = false]
  
  // Force side nodes to same rank
  {rank = same; 4; 5}
  
  // Position side nodes using invisible edges
  invisible_node [shape = point, width = 0, height = 0]
  2 -> invisible_node [style = invis]
  invisible_node -> 4 [style = invis]
  invisible_node -> 5 [style = invis]
}
") %>% 
  export_svg() %>% 
  charToRaw() %>% 
  rsvg_pdf("flowchart.pdf")

knitr::include_graphics("flowchart.pdf")

```


In the integrated Stakeholder-Legitimacy framework (@fig-flowchart), stakeholder pressures act as catalysts for banks' ESG disclosures. According to stakeholder theory, companies should consider and balance the diverse viewpoints and expectations of all groups affected by or invested in their operations, rather than focusing solely on shareholders [@BUCHHOLZ2005; @LAPLUME2008]. @FREEMAN1984 suggests that company management should stay attuned to changing dynamics and trends affecting both their organization's internal constituents and outside parties. Stakeholder pressures arise from key external actors, including regulators, investors, and customers. These actors demand greater transparency and accountability, motivating banks to provide valuable explanations of how they answer to the societal call for sustainable business conduct. Banks respond to these pressures through strategic legitimacy-seeking behaviors. Banks' ESG disclosures range from symbolic gestures (e.g. adopting TCFD guidelines without operational changes) to substantive actions such as phasing out coal financing. Crucially, legitimacy is not a static achievement but an ongoing process; banks must continuously adapt to shifting norms, such as the rise of the TNFD as a biodiversity benchmark. In the stakeholder-legitimacy context, influential stakeholders execute greater pressure on companies to explain and justify their business conduct. Therefore, sustainable reporting and disclosed ESG information serve as a way for companies to establish and maintain their legitimacy in the eyes of these stakeholders [@CAMPBELL_D2003]. 

The integrated framework further incorporates moderating contexts that shape ESG disclosures produced by the translation of stakeholder pressures into banks' legitimacy-seeking behaviors, including economic conditions such as size and profitability, and institutional context such as ownership structure and regulatory regimes. These moderators do not merely correlate with disclosure levels but fundamentally shape banks’ capacity and incentives to respond substantively or symbolically to legitimacy demands (@CAMPBELL2007; @AGUINIS2012). Below, we analyze each moderator’s role within the framework


## Moderating Context in the Stakeholder-Legitimacy Framework

### Bank (Firm) Size

Bank size (usually proxied by total assets or market capitalization) moderates the stakeholder-legitimacy dynamic by intensifying both external pressures and internal capacity. Larger frims inherently face greater visibility [@MEZNAR1995], attracting heightened scrutiny from a broader, more heterogeneous stakeholder ]base (regulators, NGOs, media) [@PELOZA2006]. Therefor, companies with significant market visibility tend to prioritize enhanced sustainability initiatives, motivated by the desire to maintain a positive reputation among their stakeholders [@DAMATO2020; @ABDI2022]. Simultaneously, size confers resource advantages: as ESG engagements involve complex processes and require a large scale to be effective, economies of scale lower the marginal cost of ESG data collection, verification, and disclosure processes [@YOUN2015; @FAVINO2019]. Consequently, larger banks are structurally positioned to deploy substantive legitimacy-seeking responses (e.g., comprehensive TCFD-aligned reporting, sector-wide decarbonization commitments) rather than symbolic gestures (@DREMPETIC2020; @LAMANDA2024). Empirical banking studies confirm this dual role, showing size positively influences ESG disclosure breadth and quality (@BUALLAY2019; @ALAREENI2020).


### Financial Performance

Financial performance (e.g., ROA, ROE) moderates the framework by determining banks’ ​resource slack—the ability to absorb ESG disclosure costs and invest in substantive legitimacy actions. A body of literature study the relation between ESG disclosure/performance and financial performance, mostly focusing on the impact of ESG activities on financial performance [@FAVINO2019; @BUALLAY2019; @DAMATO2020; @ABDI2022; @LAMANDA2024]. Financial performance as a determinant of ESG disclosure receives less academic attention. The framework posits that profitable banks possess greater operational flexibility to reallocate resources toward stakeholder-responsive ESG initiatives without jeopardizing core financial objectives [@HADDOCK2005; @LIU2009; @GAMERSCHLAG2011]. This resource slack enables more substantive legitimacy-seeking behaviours (e.g., integrating ESG risk into lending criteria, funding green bonds) aimed at genuine alignment with stakeholder norms [@AGUINIS2012]. Conversely, firms with higher leverage face heightened cost sensitivity and are less likely to make voluntary environmental disclosures due to financial constraints and costs [@BRAMMER2008].  While @DYDUCH2017 do not find a significant relationship between financial performance and sustainability reporting in Polish listed companies. @YUEN2022 report a U-shaped relationship between ESG activities and bank profitability, suggesting that only banks exceeding a threshold can translate stakeholder pressure into costly ESG actions.    


### Ownership Structure

Ownership structure and regulation can be considered as external influential factors that are beyond management control. Ownership structure critically moderates stakeholder pressures and legitimacy-seeking incentives by altering governance priorities and power asymmetries. Studies examining different ownership variables have emerged primarily in recent years. The most investigated ownership structure include a company's listing status (publicly listed), institutional shareholding, government involvement (state-owned), and shareholding concentration. Within the framework, different ownership types shape how banks prioritize which stakeholder demands merit substantive responses. 

Publicly listed companies, in order to comply with regulations and to be aligned with industry best practices, as well as to stay attuned to stakeholder pressure and see societal legitimacy, tend to be actively engaged in ESG reporting. Publicly listed companies face amplified legitimacy demands driven by regulatory compliance mandates, reputational vulnerability, and shareholder expectations. Therefore, public offering is linked to greater adoption of social reporting practices [@HADDOCK2005]; and publicly listed companies tend to disclose more sustainability-related information [@GAMERSCHLAG2011]. This compels a baseline of ESG disclosure to mitigate legitimacy threats. The publicly listed ownership structure was primarily examined as a variable in studies conducted before 2015. contemporary studies [e.g. @SAHASRANAMAM2020; @AMEEN2022; @ALOBAID2024] treat listed banks as primary disclosure agents (data source), reflecting that publicly listing embeds banks within accountability networks that amplify stakeholder salience [^04].

[^04]: Stakeholder salience refers to the degree of priority and attention that managers or organizations give to the claims and interests of different stakeholders. See @MITCHELL1997

In more recent research, institutional shareholding has emerged as a key variable of interest, such as @DYCK2019, @CHEN2020, and @AMEEN2022. ​Institutional shareholding​ transforms the stakeholder-legitimacy dynamic through active, resource-rich surveillance. Institutional investors are defined as pension funds, insurance companies, banks, sovereign wealth funds, and other institutions that manage and invest funds on behalf of their beneficiaries; they are major players in the global financial markets [@AMEEN2022]. Institutional investors, leveraging their significant equity stakes and specialized expertise, closely monitor corporate operations and pressure management, reducing agency costs and aligning operations with long-term value drivers [@SHLEIFER1986]. Due to their influential position, institutional investors can shape companies' ESG initiatives to disclose greater ESG-related data, aiming to fulfill non-financial goals and bolster both their own and the firms' public image [@DYCK2019; @GARCIA-SANCHEZ2020]. This escalation of ESG accountability demands position institutional investors as catalysts for substantive legitimacy responses. Nevertheless, empirical literature provides inconsistent findings. Some studies indicate that there is a positive association between institutional ownership and ESG performance [@DYCK2019; @CHEN2020; @AMEEN2022; @RAIMO2020]. @ZHOU2019 repots that higher institutional ownership can facilitate firms’ transparency by providing voluntary CSR reports. @ALUCHNA2022 find a negative association between institutional ownership and disclosure of the social performance. @QU2007 produces insignificant results on the relationship between corporate ownership, including institutional ownership and state ownership, and a company’s CSR engagement. These inconsistencies highlight that the framework's dependency on contex: institutional priorities (e.g., short-term returns vs. ESG integration) moderate their amplification effect on stakeholder pressures.

While institutional investors’ role in ESG practices has garnered significant attention, the influence of state ownership introduces unique legitimacy calculus, particularly in balancing public policy objectives with corporate accountability and transparency. @TAGESSON2009 find that State-owned corporations disclose more social information than privately owned corporations. One explanation is that state-owned corporations face heightened scrutiny, with pressure from their government stakeholders and public/media attention driving demands for accountability. These entities appear to have responded by aligning with stakeholder demands, reflecting compliance with external expectations: state ownership addresses market failures and improve social welfare [see @STIGLITZ1993]. The intensified social accountability often compels expansive ESG disclosure on social and development metrics by state-owned reporting entities [@TAGESSON2009; @ZHOU2019; @BOSE2018]. However, this response to social pressure is moderated by the dual legitimacy mandate. State-owned entities must reconsile commercial viability with political directives. When these confilict, ESG reporting quality may be compromised [@RAIMO2020].

Shareholding concentration disrupts stakeholder influence by enabling controlling owners to insulate legitimacy pursuits. One or several shareholders that have substantial minority ownership stakes and voting rights(such as 10 or 20 percent) diminish minority stakeholder voices through voting control and board dominance [@SHLEIFER1997]. This ​power asymmetry​ reduces external accountability pressures, permitting legitimacy-seeking behaviors that prioritize private control benefits over public transparency [@SMITH2022; @DUCASSY2015]. Controlling shareholders are less motivated to provide CSR information, with minority shareholders have less power in CSR disclosure practices [@SMITH2022; @BORGES2024]. This manifests empirically as lower CSR performance and less comprehensive sustainability reporting [@LAU2016].


### Regulatory Environment

In the Stakeholder-Legitimacy framework, the regulatory environment as an institutional moderator reshapes the translation of stakeholder pressures to banks' legitimacy-seeking disclosure. It codifies stakeholder expectations to enforceable norms, alters the cost-benefit calculus of ESG disclosure, and defines the boundaries between symbolic compliance and substantive accountability. Regulatory regimes intervene in the stakeholder-pressure mechanism by ​institutionalizing societal expectations.​​ Current voluntary initialtives of social reporting are insufficienct to achieve coporate accountability [@HESS2007]. When regulators formalise ESG disclosure requirements (e.g., EU CSRD, national green finance frameworks), they elevate specific stakeholder concerns to mandated accountability obligations [@QU2007]. This ​compression of stakeholder pressures​ replaces fragmented market demands with codified legitimacy thresholds. Consequently, regulatory stringency strengthens both the rigor and reliability of corporate disclosure practices, fostering greater societal accountability [@CICCHIELLO2023]. The role of regulatory environment is further acknowledged in ​reconfiguring information asymmetries and signalling costs. ​@DESAI2024 reports that stock market investors reacts positively to the mandatory ESG disclosure regulation because mandatory ESG disclosure is considered as a positive signal by reducing information asymmetry. The compliance threshold deiminishes the legitimacy benefits of superficial compliance. When ESG disclosure metrics become audited and standardised, the signalling value of substantive reporting intensifites while symbolic gestures risk regulatory penalties and reputational damage [@DESAI2024; @BOSE2018].

Furthermore, regulatory heterogeneity across jurisdictions ​moderates stakeholder pressure translation.​​ Banks in weak institutional environments may exploit voluntary regimes to pursue legitimacy through ceremonial conformance, adopting frameworks like PRB while decoupling operations from disclosures [@MANOS2024]. Conversely, stringent regulations compel operational restructuring. Studies of green banking mandates demonstrate tighter coupling between policy requirements and substantive environmental risk integration [@BOSE2018]. This contingency underscores the framework’s emphasis on regulatory quality in determining whether stakeholder pressures catalyse authentic accountability or reputational arbitrage.

This section presents an integrated Stakeholder-Legitimacy Framework, tailored to the banking sector, to explain banks’ ESG disclosure practices, synthesizing stakeholder theory, legitimacy theory, and institutional contexts into a cohesive analytical pathway. The framework posits that ESG disclosures arise from dynamic interactions between external stakeholder pressures (e.g., regulators, investors, customers) and banks’ legitimacy-seeking behaviors, moderated by economic factors (e.g., size, profitability) and institutional conditions (e.g., ownership structure, regulatory regimes). By bridging insights from @CAMPBELL2007 and @AGUINIS2012, the framework advances a holistic view of ESG reporting as a negotiated outcome of strategic adaptation to stakeholder demands and institutional expectations.


# Conclusion

The transition from voluntary ESG initiatives to mandatory disclosure frameworks marks a paradigm shift in how financial institutions conceptualize and manage nature-related risks. Global frameworks like the TNFD and ISSB, alongside regional regulations such as the EU Taxonomy and CSRD, have established rigorous reporting standards that integrate environmental impacts into financial decision-making. For banks, these mandates necessitate not only transparency about their own operations but also heightened scrutiny of borrowers’ sustainability practices [see @WANG2023], creating cascading effects across economies. Empirical evidence suggests that ESG disclosures can mitigate bank-specific risks—such as insolvency and liquidity challenges—while improving market efficiency through reduced information asymmetry [@GANGI2019; @SCHOLTENS2019; @GANGWANI2024]. However, conflicting findings on the financial performance-ESG relationship and the role of ownership structures underscore the need for contextualized analyses.

To address these complexities, this study advances the Stakeholder-Legitimacy Framework and tailor it to the banking sector. This framework synthesizes stakeholder theory, legitimacy theory, and institutional isomorphism to explain how banks’ ESG disclosures emerge from dynamic negotiations between external pressures (e.g., regulators, investors) and legitimacy-seeking behaviors. It posits that banks balance competing stakeholder demands—such as investor calls for transparency and regulatory mandates for biodiversity due diligence—through strategic disclosures that range from symbolic compliance (e.g., superficial TNFD adoption) to substantive action (e.g., divesting from deforestation-linked assets). Crucially, the framework incorporates moderating factors such as ownership structures (e.g., institutional ownership, state ownership) and regulatory regimes (e.g., mandatory CSRD, voluntary IFRS S1 and S2 adoption), offering a nuanced lens to analyze how economic and institutional contexts shape disclosure outcomes [@CAMPBELL2007; @AGUINIS2012]. 

Critical research gaps persist, however. First, while institutional and state ownership structures are increasingly scrutinized, findings remain inconsistent: institutional ownership’s role in ESG performance varies across studies [e.g., @DYCK2019; @ALUCHNA2022], and state ownership’s dual mandate for social welfare and corporate accountability lacks consensus [@TAGESSON2009; @RAIMO2020]. Second, the moderating role of regulatory frameworks is underexplored, particularly in the highly regulated banking sector. Empirical research on how mandatory ESG regulations (e.g., CSRD, EU Taxonomy) shape bank risk profiles, financial performance, and governance mechanisms remains scarce, despite the systemic importance of banks and their unique exposure to ESG-related risks. While studies have begun analyzing disclosure quality versus symbolic compliance in non-financial sectors [@CICCHIELLO2023; @MANOS2024], sector-specific analyses for banking—where regulatory pressures intersect with complex risk intermediation roles—are urgently needed. Third, compounding this complexity, banks’ dual role in ESG ecosystems warrants deeper scrutiny: as both reporting entities subject to disclosure mandates and financing intermediaries influencing firms’ access to green capital, they create transmission effects that amplify regulatory impacts across economies. For instance, @WANG2023 demonstrates that ESG disclosure regulations propagate through bank lending networks, altering borrowers’ environmental practices even without direct regulatory coverage—a phenomenon underexplored in the literature. Fourth, mixed evidence on the relationship between financial performance and ESG disclosure—such as conflicting results on profitability [@HADDOCK2005; @DYDUCH2017]—highlights the need for contextualized analyses. Finally, concentrated ownership’s tendency to marginalize minority shareholder influence on CSR practices [@SMITH2022; @BORGES2024] underscores a governance gap in aligning dominant shareholders with broader stakeholder expectations. These gaps signal opportunities for future research to refine the framework, particularly in disentangling ownership dynamics, regulatory efficacy, and the interplay between symbolic and substantive ESG strategies across diverse institutional settings.

Future research may prioritize three avenues: (1) disentangling the economic and institutional factors that drive ESG disclosure quality in banks; (2) evaluating the systemic risk implications of nature-related regulatory shocks; for instance, assess how nature-related disclosures influence systemic risk within the banking sector and the wider economy; and (3) exploring the interaction between stakeholder-legitimacy framework and institutional isomorphism framework, to assess how coercive, mimetic, and normative pressures shape banks' ESG disclosure practices. 

The proliferation of nature-related disclosure frameworks in recent years reveals the shift from vulentary toward mandatory disclosure: from TNFD's market-driven recommendations to IFRS S1 and S2 disclosure framework, and mandatory EU reporting standards CSRD. This trend is driven by several key developments including recognition of material nature-related risks, global reporting alignment, and integration with financial reporting. However, accelerating fragmentation creates compliance complexity, while implementation gaps persist in emerging economies facing data scarcity and capacity constraints. The fourth research direction may therefore focus on: harmonization pathways across frameworks, effectiveness of disclosure mandates in redirecting conservation financing (e.g. how metric inconsistencies create greenwashing vulnerabilities), and tangible impacts on biodiversity loss reduction—particularly in high-risk lending portfolios.​

By focusing on these research avenues through the lens of the Stakeholder-Legitimacy Framework, scholars and policymakers can enhance our understanding of ESG disclosure dynamics in the banking sector. This will not only refine regulatory architectures but also ensure that ESG disclosures go beyond compliance checklists to drive meaningful progress. 


# Reference


























